---
title: "Kenya MESS"
format: html
---

## Setup

```{r}
#| label: setup
#| message: false

library(targets)
library(terra)
library(dplyr)
library(tidyr)
library(ggplot2)
library(bssdm)

tar_load(c(covariate_rast_5, model_data_spatial, ken, target_covariate_names))

# Muted divergent shades used throughout (negative/novel vs positive/similar)
col_neg <- "#D6604D"
col_pos <- "#4393C3"
```

## Crop and mask raster to Kenya

```{r}
#| label: cov-rast-ken

cov_rast_ken <- crop(covariate_rast_5, ken) |>
  mask(ken)
```

## Create SpatVector from model data

```{r}
#| label: model-data-vect

model_data_vect <- vect(
  model_data_spatial |>
    select(
      species,
      longitude,
      latitude,
      count,
      presence,
      data_type
    ),
  geom = c("longitude", "latitude"),
  crs = crs(ken)
)
```

## Subset points within Kenya

```{r}
#| label: ken-points

ken_points <- model_data_vect[ken, ]
```

## MESS iteration 0 (before adding any points)

Subset the raster to the target covariate layers and, for each species, compute
the reference environmental values from its observation points and the initial
(iteration 0) MESS surface -- i.e. the MESS before any search points are added.

```{r}
#| label: mess-iter0

cov_rast_ken_sub <- cov_rast_ken[[target_covariate_names]]

target_species <- c("arabiensis", "gambiae", "funestus")

# Reference env values per species (observation points only)
ref_vals_0 <- lapply(target_species, function(sp) {
  sp_points <- ken_points[ken_points$species == sp, ]
  terra::extract(cov_rast_ken_sub, sp_points, ID = FALSE) |>
    na.omit()
})
names(ref_vals_0) <- target_species

# Iteration 0 MESS surface per species
mess_iter0 <- lapply(ref_vals_0, function(rv) mess(cov_rast_ken_sub, rv))
names(mess_iter0) <- target_species
```

### Iteration 0 MESS by species

```{r}
#| label: plot-iter0
#| fig-width: 12
#| fig-height: 5

iter0 <- lapply(mess_iter0, \(m) m$mess)

# Shared symmetric range, clamped to the 1st/99th percentile so extreme outliers
# don't wash out the colour variation. Zero maps to white for comparability.
iter0_absmax <- max(abs(unlist(lapply(iter0, \(r) {
  quantile(values(r), c(0.01, 0.99), na.rm = TRUE)
}))))
iter0_range  <- c(-iter0_absmax, iter0_absmax)
divergent_pal <- colorRampPalette(c(col_neg, "white", col_pos))(255)

par(mfrow = c(1, 3), mar = c(2, 2, 3, 4))
for (sp in target_species) {
  plot(iter0[[sp]], main = sp, col = divergent_pal, range = iter0_range, axes = FALSE)
}
```

### Pairwise differences between species (iteration 0)

Each panel maps the cell-by-cell difference in iteration 0 MESS between a pair of
species. A shared symmetric scale is used so zero (no difference) maps to white.

```{r}
#| label: plot-iter0-diff
#| fig-width: 12
#| fig-height: 5

iter0_diffs <- list(
  "arabiensis - gambiae"  = iter0$arabiensis - iter0$gambiae,
  "arabiensis - funestus" = iter0$arabiensis - iter0$funestus,
  "gambiae - funestus"    = iter0$gambiae    - iter0$funestus
)

diff_absmax <- max(abs(unlist(lapply(iter0_diffs, \(r) minmax(r)))))
diff_range  <- c(-diff_absmax, diff_absmax)

# Distinct divergent palette (orange-teal) for the difference maps, symmetric
# around zero so no difference maps to white.
diff_pal <- colorRampPalette(c("#E08214", "white", "#80CDC1"))(255)

# Layout: three map panels above one shared horizontal colourbar
layout(rbind(1:3, c(4, 4, 4)), heights = c(1, 0.28))
par(mar = c(2, 2, 3, 1))
for (nm in names(iter0_diffs)) {
  plot(iter0_diffs[[nm]], main = nm, col = diff_pal, range = diff_range,
       legend = FALSE, axes = FALSE)
}

# Shared colourbar
par(mar = c(3, 6, 0.5, 6))
diff_vals <- seq(diff_range[1], diff_range[2], length.out = length(diff_pal))
image(x = diff_vals, y = 1, z = matrix(diff_vals, ncol = 1),
      col = diff_pal, axes = FALSE, xlab = "Difference in iteration 0 MESS", ylab = "")
axis(1)
```

### Novel (negative) regions -- are they similar across species?

Mask each iteration 0 surface to keep only the negative (novel / extrapolation)
cells, then compare where the novel regions fall across species.

```{r}
#| label: iter0-negative-similarity

# Keep only negative cells (novel areas)
iter0_neg <- lapply(iter0, \(r) { r[r >= 0] <- NA; r })

# Which cells are novel (negative) per species?
is_neg <- lapply(iter0, \(r) values(r) < 0)

# Pairwise agreement in the *location* of novel cells (Jaccard-style overlap)
overlap <- function(a, b) {
  both <- a & b
  either <- a | b
  sum(both, na.rm = TRUE) / sum(either, na.rm = TRUE)
}

cat("Overlap of novel-cell locations (intersection / union):\n",
  "arabiensis vs gambiae: ", round(overlap(is_neg$arabiensis, is_neg$gambiae),  3), "\n",
  "arabiensis vs funestus:", round(overlap(is_neg$arabiensis, is_neg$funestus), 3), "\n",
  "gambiae vs funestus:   ", round(overlap(is_neg$gambiae,    is_neg$funestus), 3), "\n"
)

# Correlation of the negative MESS values where both species are novel
neg_cor <- function(a, b) {
  va <- values(a)[, 1]; vb <- values(b)[, 1]
  ok <- !is.na(va) & !is.na(vb)
  if (sum(ok) < 2) return(NA_real_)
  cor(va[ok], vb[ok])
}

cat("\nCorrelation of negative MESS values where both are novel:\n",
  "arabiensis vs gambiae: ", round(neg_cor(iter0_neg$arabiensis, iter0_neg$gambiae),  3), "\n",
  "arabiensis vs funestus:", round(neg_cor(iter0_neg$arabiensis, iter0_neg$funestus), 3), "\n",
  "gambiae vs funestus:   ", round(neg_cor(iter0_neg$gambiae,    iter0_neg$funestus), 3), "\n"
)
```

## MESS surfaces and iterative novel-environment point search

For each of the three target species, iteratively generate a MESS surface
(starting from the iteration 0 surface above), identify the cell with the lowest
similarity score, record its centroid, add it to the reference set, and repeat
until 20 points have been recorded.

```{r}
#| label: mess-iteration

n_iter <- 20
mess_results <- list()

for (sp in target_species) {

  # Start from the species observation points
  ref_vals <- ref_vals_0[[sp]]

  recorded_pts <- data.frame(longitude = numeric(0), latitude = numeric(0))
  # Surfaces indexed 1..(n_iter + 1) for iterations 0..n_iter; iteration 0 is
  # the MESS before any search points have been added
  mess_surfaces <- vector("list", n_iter + 1)

  for (i in 0:n_iter) {

    # Reuse the pre-computed iteration 0 surface; otherwise recompute
    m <- if (i == 0) mess_iter0[[sp]] else mess(cov_rast_ken_sub, ref_vals)
    mess_surfaces[[i + 1]] <- m

    # Final surface computed; no further points to add
    if (i == n_iter) break

    # Identify cell with lowest MESS value and get its centroid
    min_cell <- which.min(values(m$mess))
    centroid  <- xyFromCell(m$mess, min_cell)
    lon <- centroid[1, "x"]
    lat <- centroid[1, "y"]

    recorded_pts <- rbind(
      recorded_pts,
      data.frame(longitude = lon, latitude = lat)
    )

    # Extract env values at new point and add to reference set
    new_pt   <- vect(data.frame(x = lon, y = lat), geom = c("x", "y"), crs = crs(cov_rast_ken_sub))
    new_vals <- terra::extract(cov_rast_ken_sub, new_pt, ID = FALSE)
    ref_vals <- rbind(ref_vals, new_vals)
  }

  mess_results[[sp]] <- list(
    mess_surfaces = mess_surfaces,
    recorded_points = recorded_pts
  )

  message("Done: ", sp)
}
```

## Results

```{r}
#| label: results

for (sp in target_species) {
  cat("\n###", sp, "\n\n")
  print(mess_results[[sp]]$recorded_points)
}
```


```{r}
#| label: selected-points-identical

# Check whether the selected points are identical across the three species
selected_pts <- lapply(mess_results, \(x) x$recorded_points)

cat(
  "arabiensis vs gambiae: ", identical(selected_pts$arabiensis, selected_pts$gambiae),  "\n",
  "arabiensis vs funestus:", identical(selected_pts$arabiensis, selected_pts$funestus), "\n",
  "gambiae vs funestus:   ", identical(selected_pts$gambiae,    selected_pts$funestus), "\n",
  "all three identical:   ",
  identical(selected_pts$arabiensis, selected_pts$gambiae) &&
    identical(selected_pts$arabiensis, selected_pts$funestus), "\n"
)
```



## Are species observation points identical?

```{r}
#| label: points-identity

coords_by_species <- lapply(target_species, function(sp) {
  pts <- crds(ken_points[ken_points$species == sp, ])
  sort(paste(round(pts[, "x"], 6), round(pts[, "y"], 6)))
})
names(coords_by_species) <- target_species

# Point counts per species
sapply(coords_by_species, length)

# Pairwise identity checks
cat(
  "arabiensis vs gambiae: ", identical(coords_by_species[["arabiensis"]], coords_by_species[["gambiae"]]), "\n",
  "arabiensis vs funestus:", identical(coords_by_species[["arabiensis"]], coords_by_species[["funestus"]]), "\n",
  "gambiae vs funestus:   ", identical(coords_by_species[["gambiae"]],    coords_by_species[["funestus"]]),  "\n"
)
```

## Absolute sum of positive and negative cells per iteration

For each surface (from iteration 0), sum the positive cell values (similar,
blue) and the absolute value of the negative cells (novel, red), then track how
these totals change across iterations. The grey line (secondary axis) shows the
MESS value of the point selected at each iteration -- i.e. how novel the most
novel cell is -- so we can see what is being picked as the search continues.

```{r}
#| label: mess-cell-sums
#| fig-width: 12
#| fig-height: 4

cell_sums <- lapply(target_species, function(sp) {
  surfaces <- mess_results[[sp]]$mess_surfaces
  data.frame(
    species   = sp,
    iteration = seq_along(surfaces) - 1,
    positive  = sapply(surfaces, function(m) {
      v <- values(m$mess); sum(v[v > 0], na.rm = TRUE)
    }),
    negative  = sapply(surfaces, function(m) {
      v <- values(m$mess); sum(abs(v[v < 0]), na.rm = TRUE)
    })
  )
}) |>
  bind_rows() |>
  pivot_longer(c(positive, negative), names_to = "sign", values_to = "abs_sum")

# MESS value of the point selected at each iteration (the most novel cell).
# Points are selected at iterations 0..(n_iter - 1); the final surface has none.
selected_mess <- lapply(target_species, function(sp) {
  surfaces <- mess_results[[sp]]$mess_surfaces
  data.frame(
    species    = sp,
    iteration  = seq_len(n_iter) - 1,
    mess_value = sapply(surfaces[seq_len(n_iter)], function(m) {
      min(values(m$mess), na.rm = TRUE)
    })
  )
}) |>
  bind_rows()

# Combine into one long frame with a panel per metric: absolute sums on the top
# row, selected-point MESS value on its own row below (each with a free y-axis).
plot_data <- bind_rows(
  cell_sums |>
    transmute(species, iteration, series = sign, value = abs_sum,
              panel = "Absolute sum of MESS cells"),
  selected_mess |>
    transmute(species, iteration, series = "selected", value = mess_value,
              panel = "Selected point MESS value")
) |>
  mutate(panel = factor(
    panel,
    levels = c("Absolute sum of MESS cells", "Selected point MESS value")
  ))

ggplot(plot_data, aes(iteration, value, colour = series)) +
  geom_line(data = ~ filter(.x, series == "selected")) +
  geom_point() +
  facet_grid(panel ~ species, scales = "free_y", switch = "y") +
  scale_colour_manual(
    values = c(negative = col_neg, positive = col_pos, selected = "grey40"),
    labels = c(
      negative = "|negative| (novel)",
      positive = "positive (similar)",
      selected = "selected point MESS"
    ),
    breaks = c("negative", "positive", "selected")
  ) +
  labs(x = "Iteration", y = NULL, colour = NULL) +
  theme(strip.placement = "outside")
```

## Absolute sum table

```{r}
#| label: mess-cell-sums-table

cell_sums |>
  pivot_wider(names_from = sign, values_from = abs_sum) |>
  left_join(selected_mess, by = c("species", "iteration")) |>
  arrange(species, iteration) |>
  knitr::kable(
    digits    = c(0, 0, 0, 0, 2),
    col.names = c("Species", "Iteration", "positive (similar)", "|negative| (novel)", "Selected point MESS")
  )
```

## MESS surface evolution by species

Each panel shows the MESS surface at that iteration (starting from iteration 0,
before any points are added), with all points selected up to and including that
iteration overlaid in black.

```{r}
#| label: plot-mess
#| fig-width: 10
#| fig-height: 11

# Clamp to the 1st/99th percentile across all surfaces (symmetric around zero)
# so extreme negative outliers don't wash out the colour variation. Zero maps
# to white and all panels share the same scale for comparability.
clamp_absmax <- max(abs(unlist(lapply(target_species, function(sp) {
  lapply(mess_results[[sp]]$mess_surfaces, function(m) {
    quantile(values(m$mess), c(0.01, 0.99), na.rm = TRUE)
  })
}))))

mess_range <- c(-clamp_absmax, clamp_absmax)

# Divergent palette: muted red = negative (novel), white = zero, blue = positive
divergent_pal <- colorRampPalette(c(col_neg, "white", col_pos))(255)

# Layout: 3x3 grid of panels plus a bottom strip for the shared legend
layout_mat <- rbind(
  matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE),
  rep(10, 3)
)

for (sp in target_species) {
  pts      <- mess_results[[sp]]$recorded_points
  surfaces <- mess_results[[sp]]$mess_surfaces

  layout(layout_mat, heights = c(1, 1, 1, 0.4))
  par(mar = c(1, 1, 2, 0.5), oma = c(0, 0, 3, 0))

  # Iterations 0 through 8 (surfaces 1 through 9)
  for (j in seq_len(9)) {
    iter <- j - 1
    plot(
      surfaces[[j]]$mess,
      main   = paste0("iter ", iter),
      col    = divergent_pal,
      range  = mess_range,
      legend = FALSE,
      axes   = FALSE
    )
    lines(ken)

    # Show all points selected up to and including this iteration
    if (iter >= 1) {
      pts_so_far <- pts[seq_len(iter), , drop = FALSE]
      pts_vect   <- vect(pts_so_far, geom = c("longitude", "latitude"), crs = crs(cov_rast_ken_sub))
      points(pts_vect, pch = 19, col = "goldenrod", cex = 2)
    }
  }

  # Shared legend (panel 10): horizontal colour bar spanning the grid width
  par(mar = c(2.5, 3, 0.5, 3))
  legend_vals <- seq(mess_range[1], mess_range[2], length.out = length(divergent_pal))
  image(
    x = legend_vals, y = 1, z = matrix(legend_vals, ncol = 1),
    col = divergent_pal, axes = FALSE, xlab = "MESS (clamped to 1st/99th pctile)", ylab = ""
  )
  axis(1)

  mtext(sp, outer = TRUE, cex = 1.4, font = 2)
}
```

# Funestus-driven point selection

Here the iterative search selects points **only** from the funestus MESS surface,
but each selected point is added to *all three* species' reference sets. This
shows how points chosen to fill funestus's novel environment reshape the MESS of
the other two species as well.

```{r}
#| label: mess-iteration-funestus

# Reference sets start from each species' observation points
ref_vals_fun <- ref_vals_0

recorded_pts_fun   <- data.frame(longitude = numeric(0), latitude = numeric(0))
selected_cells_fun <- integer(0)
mess_surfaces_fun  <- setNames(
  lapply(target_species, \(sp) vector("list", n_iter + 1)),
  target_species
)

for (i in 0:n_iter) {

  # MESS for every species with the current (shared) reference sets
  for (sp in target_species) {
    mess_surfaces_fun[[sp]][[i + 1]] <-
      if (i == 0) mess_iter0[[sp]] else mess(cov_rast_ken_sub, ref_vals_fun[[sp]])
  }

  if (i == n_iter) break

  # Select the most novel cell from the FUNESTUS surface only
  m_fun    <- mess_surfaces_fun[["funestus"]][[i + 1]]
  min_cell <- which.min(values(m_fun$mess))
  centroid <- xyFromCell(m_fun$mess, min_cell)
  lon <- centroid[1, "x"]
  lat <- centroid[1, "y"]

  recorded_pts_fun   <- rbind(recorded_pts_fun, data.frame(longitude = lon, latitude = lat))
  selected_cells_fun <- c(selected_cells_fun, min_cell)

  # Add the funestus-selected point to ALL species' reference sets
  new_pt   <- vect(data.frame(x = lon, y = lat), geom = c("x", "y"), crs = crs(cov_rast_ken_sub))
  new_vals <- terra::extract(cov_rast_ken_sub, new_pt, ID = FALSE)
  for (sp in target_species) ref_vals_fun[[sp]] <- rbind(ref_vals_fun[[sp]], new_vals)
}

# Package like mess_results; recorded points are shared (the funestus picks)
mess_results_fun <- setNames(
  lapply(target_species, \(sp) list(
    mess_surfaces   = mess_surfaces_fun[[sp]],
    recorded_points = recorded_pts_fun
  )),
  target_species
)
```

## Absolute sum of positive and negative cells per iteration (funestus-driven)

The selected-point MESS series now shows the value of the *funestus-selected*
cell on each species' own surface -- for funestus this is its minimum, for the
others it is how novel that same point is to them.

```{r}
#| label: mess-cell-sums-funestus
#| fig-width: 12
#| fig-height: 4

cell_sums_fun <- lapply(target_species, function(sp) {
  surfaces <- mess_results_fun[[sp]]$mess_surfaces
  data.frame(
    species   = sp,
    iteration = seq_along(surfaces) - 1,
    positive  = sapply(surfaces, function(m) {
      v <- values(m$mess); sum(v[v > 0], na.rm = TRUE)
    }),
    negative  = sapply(surfaces, function(m) {
      v <- values(m$mess); sum(abs(v[v < 0]), na.rm = TRUE)
    })
  )
}) |>
  bind_rows() |>
  pivot_longer(c(positive, negative), names_to = "sign", values_to = "abs_sum")

# MESS value at the funestus-selected cell, evaluated on each species' surface
selected_mess_fun <- lapply(target_species, function(sp) {
  surfaces <- mess_results_fun[[sp]]$mess_surfaces
  data.frame(
    species    = sp,
    iteration  = seq_len(n_iter) - 1,
    mess_value = sapply(seq_len(n_iter), function(i) {
      values(surfaces[[i]]$mess)[selected_cells_fun[i]]
    })
  )
}) |>
  bind_rows()

plot_data_fun <- bind_rows(
  cell_sums_fun |>
    transmute(species, iteration, series = sign, value = abs_sum,
              panel = "Absolute sum of MESS cells"),
  selected_mess_fun |>
    transmute(species, iteration, series = "selected", value = mess_value,
              panel = "Selected point MESS value")
) |>
  mutate(panel = factor(
    panel,
    levels = c("Absolute sum of MESS cells", "Selected point MESS value")
  ))

ggplot(plot_data_fun, aes(iteration, value, colour = series)) +
  geom_line(data = ~ filter(.x, series == "selected")) +
  geom_point() +
  facet_grid(panel ~ species, scales = "free_y", switch = "y") +
  scale_colour_manual(
    values = c(negative = col_neg, positive = col_pos, selected = "grey40"),
    labels = c(
      negative = "|negative| (novel)",
      positive = "positive (similar)",
      selected = "selected point MESS"
    ),
    breaks = c("negative", "positive", "selected")
  ) +
  labs(x = "Iteration", y = NULL, colour = NULL) +
  theme(strip.placement = "outside")
```

## Absolute sum table (funestus-driven)

```{r}
#| label: mess-cell-sums-table-funestus

cell_sums_fun |>
  pivot_wider(names_from = sign, values_from = abs_sum) |>
  left_join(selected_mess_fun, by = c("species", "iteration")) |>
  arrange(species, iteration) |>
  knitr::kable(
    digits    = c(0, 0, 0, 0, 2),
    col.names = c("Species", "Iteration", "positive (similar)", "|negative| (novel)", "Selected point MESS")
  )
```

## MESS surface evolution by species (funestus-driven)

Each panel shows the MESS surface at that iteration. The overlaid points are the
funestus-selected locations, identical across all three species' panels.

```{r}
#| label: plot-mess-funestus
#| fig-width: 10
#| fig-height: 11

clamp_absmax_fun <- max(abs(unlist(lapply(target_species, function(sp) {
  lapply(mess_results_fun[[sp]]$mess_surfaces, function(m) {
    quantile(values(m$mess), c(0.01, 0.99), na.rm = TRUE)
  })
}))))

mess_range_fun <- c(-clamp_absmax_fun, clamp_absmax_fun)

for (sp in target_species) {
  pts      <- mess_results_fun[[sp]]$recorded_points
  surfaces <- mess_results_fun[[sp]]$mess_surfaces

  layout(layout_mat, heights = c(1, 1, 1, 0.4))
  par(mar = c(1, 1, 2, 0.5), oma = c(0, 0, 3, 0))

  for (j in seq_len(9)) {
    iter <- j - 1
    plot(
      surfaces[[j]]$mess,
      main   = paste0("iter ", iter),
      col    = divergent_pal,
      range  = mess_range_fun,
      legend = FALSE,
      axes   = FALSE
    )
    lines(ken)

    if (iter >= 1) {
      pts_so_far <- pts[seq_len(iter), , drop = FALSE]
      pts_vect   <- vect(pts_so_far, geom = c("longitude", "latitude"), crs = crs(cov_rast_ken_sub))
      points(pts_vect, pch = 19, col = "goldenrod", cex = 2)
    }
  }

  par(mar = c(2.5, 3, 0.5, 3))
  legend_vals <- seq(mess_range_fun[1], mess_range_fun[2], length.out = length(divergent_pal))
  image(
    x = legend_vals, y = 1, z = matrix(legend_vals, ncol = 1),
    col = divergent_pal, axes = FALSE, xlab = "MESS (clamped to 1st/99th pctile)", ylab = ""
  )
  axis(1)

  mtext(paste0(sp, " (funestus-driven picks)"), outer = TRUE, cex = 1.4, font = 2)
}
```
