---
title: "*Anopheles arabiensis* environmental novelty"
subtitle: "MESS and ExDet, whole of Africa and masked to the core expert range"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-align: center
execute:
  warning: false
  message: false
  echo: false
fig-width: 8
fig-height: 8
---

```{r}
#| label: setup

library(targets.utils) # pak::pak("geryan/targets.utils")
library(bssdm)         # remotes::install_github("cebra-analytics/bssdm")
library(tmap)

# quarto renders with the working directory set to this file's directory
archive_dir <- "."

read_novelty <- function(object_name, suffix) {
  read_terra_nested(
    file.path(archive_dir, paste0(object_name, suffix, ".tar.gz"))
  )
}
```


{{< pagebreak >}}

Two measures of environmental novelty: the Multivariate Environmental
Similarity Surface, MESS
([Elith et al. 2010](https://doi.org/10.1111/j.2041-210X.2010.00036.x)), and
Extrapolation Detection, ExDet
([Mesgaran et al. 2014](https://doi.org/10.1111/ddi.12209)).

| Measure | Value | Colour | Interpretation |
|:--------|:------|:-------|:-----------------------------------------------------|
| MESS    | > 0   | blue   | Within the univariate range of all variables |
| MESS    | ~ 0   | white  | Within range, but near the edge of one or more variables |
| MESS    | < 0   | red    | Extrapolating |
| ExDet   | 0 to 1 | yellow | Similar: within the multivariate / combinatorial range of the variables |
| ExDet   | < 0   | red    | Type 1 novelty: true extrapolation, the same thing as MESS < 0 |
| ExDet   | > 1   | blue   | Type 2 novelty: combinatorial extrapolation. Variables are each within their univariate range, but in unseen combinations |


# Whole of Africa

## Excluding proximity to sea

### MESS
```{r}
#| label: mess-nosea-unmasked
x <- read_novelty("africa_mess_nosea_arabiensis", "")
plot(x, which = "mess")
plot(x, which = "mod")
plot(x, which = "mos")
```


### ExDet

```{r}
#| label: exdet-nosea-unmasked
x <- read_novelty("africa_exdet_nosea_arabiensis", "")
plot(x, which = "exdet")
```


## All covariates

### MESS

```{r}
#| label: mess-unmasked
x <- read_novelty("africa_mess_arabiensis", "")
plot(x, which = "mess")
plot(x, which = "mod")
plot(x, which = "mos")
```

### ExDet

```{r}
#| label: exdet-unmasked
x <- read_novelty("africa_exdet_arabiensis", "")
plot(x, which = "exdet")
```


# Masked to the core arabiensis expert range

## Excluding prox_to_sea

### MESS

```{r}
#| label: mess-nosea-masked
x <- read_novelty("africa_mess_nosea_arabiensis", "_masked")
plot(x, which = "mess")
plot(x, which = "mod")
plot(x, which = "mos")
```


### ExDet

```{r}
#| label: exdet-nosea-masked
x <- read_novelty("africa_exdet_nosea_arabiensis", "_masked")
plot(x, which = "exdet")
```

## All covariates

### MESS 

```{r}
#| label: mess-masked
x <- read_novelty("africa_mess_arabiensis", "_masked")
plot(x, which = "mess")
plot(x, which = "mod")
plot(x, which = "mos")
```

### ExDet

```{r}
#| label: exdet-masked
x <- read_novelty("africa_exdet_arabiensis", "_masked")
plot(x, which = "exdet")
```



