---
title: "Selecting 200 survey points for *Anopheles* in Kenya"
subtitle: "Uncertainty-weighted novelty sampling with MESS, validated with ExDet"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "code"
    fig-align: center
    theme: cosmo
execute:
  warning: false
  message: false
---

# Summary

Large parts of Kenya have environments that are poorly represented by the
existing *Anopheles* occurrence records, so species-distribution predictions
there are extrapolated and uncertain. This report selects **200 new survey
locations** that best fill that gap for three vectors — *An. arabiensis*,
*An. gambiae* and *An. funestus* — with a greedy, uncertainty-weighted
novelty-sampling scheme:

- **Novelty** is measured with **MESS** (the Multivariate Environmental
  Similarity Surface): how far each pixel's environment falls outside the
  covariate ranges seen at existing records.
- **Uncertainty** is the per-pixel **coefficient of variation (CV)** of the
  current distribution models.
- Each iteration adds the single **most-novel × most-uncertain** unsampled
  pixel, *shared* across all three species, then recomputes — so the whole set
  is built around common points.

The 200 points, in priority order, are exported to
`k_m_r_selected_points.csv`. An independent metric (**ExDet**) confirms the
design eliminates *univariate* environmental novelty within the first ~16 points
and then steadily reduces the harder *combinational* novelty.

```{r}
#| label: setup

library(targets)
library(terra)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(bssdm)

# ---- Spatial inputs (targets store) ----------------------------------------
tar_load(
  c(covariate_rast_5, model_data_spatial, ken, target_covariate_names),
  store = "~/Documents/tki_work/vector_atlas/va_multispecies_sdm/_targets/"
)

# Covariate stack cropped/masked to Kenya, reduced to the modelling covariates.
cov_rast_ken_sub <- crop(covariate_rast_5, ken) |>
  mask(ken) |>
  subset(target_covariate_names)

# Species and the representative species used for the per-species evolution
# grids (the three species' novelty patterns are near-identical, see below).
ken_mess_spp <- c("arabiensis", "gambiae", "funestus")
rep_sp       <- "gambiae"

# ---- Model uncertainty: per-species coefficient of variation ----------------
cv <- rast(paste0(
  "~/Documents/tki_work/vector_atlas/va_multispecies_sdm/",
  "spartan_model_comparison/m6/m6_mask_p_cv.tif"
)) |>
  crop(ken) |>
  mask(ken)
cv <- cv[[ken_mess_spp]]
cv <- resample(cv, cov_rast_ken_sub, method = "bilinear")   # align to covariates

# ---- Existing occurrence points within Kenya, per species -------------------
ken_points <- vect(
  model_data_spatial |>
    select(species, longitude, latitude) |>
    filter(species %in% ken_mess_spp),
  geom = c("longitude", "latitude"), crs = crs(ken)
)
ken_points <- ken_points[ken, ]

# Reference environmental values per species (their observation points only).
ref_vals_0 <- setNames(
  lapply(ken_mess_spp, \(sp) {
    terra::extract(cov_rast_ken_sub, ken_points[ken_points$species == sp, ],
                   ID = FALSE) |> na.omit()
  }),
  ken_mess_spp
)
```

```{r}
#| label: helpers

# Palette / colours reused throughout. Red = novel, blue = similar.
col_neg <- "#D6604D"
col_pos <- "#4393C3"
divergent_pal <- colorRampPalette(c(col_neg, "white", col_pos))(255)

# 3x3 map grid plus a colourbar/legend strip below it.
layout_mat <- rbind(matrix(1:9, nrow = 3, byrow = TRUE), rep(10, 3))

# Symmetric colour range clamped to the 1st/99th percentile of a list of
# rasters, so outliers don't wash out the variation.
clamp_range <- function(rasters) {
  absmax <- max(abs(unlist(lapply(
    rasters, \(r) quantile(values(r), c(0.01, 0.99), na.rm = TRUE)
  ))))
  c(-absmax, absmax)
}

# Horizontal colourbar spanning a panel.
colorbar_h <- function(rng, label, pal = divergent_pal) {
  vals <- seq(rng[1], rng[2], length.out = length(pal))
  image(x = vals, y = 1, z = matrix(vals, ncol = 1), col = pal,
        axes = FALSE, xlab = label, ylab = "")
  axis(1)
}

# ExDet is shown with deterministic discrete classes (fixed `breaks`), NOT a
# continuous palette: terra re-interpolates a long `col` vector and bleeds the
# near-zero colour into the most *typical* cells, which multiply as points are
# added and look misleadingly like growing novelty. Fixed breaks keep the
# [0, 1] "similar" band grey. Lower "similar" bound is -1e-9 so exdet == 0 is
# grey, not the reddest class.
exdet_breaks <- c(-Inf, -1, -0.25, -1e-9, 1, 2, 4, Inf)
exdet_cols   <- c("#67001F", "#D6604D", "#FDDBC7",   # Type 1: strong -> weak
                  "#f0f0f0",                          # similar
                  "#E7D4E8", "#9970AB", "#40004B")    # Type 2: weak -> strong
exdet_class_labels <- c(
  "Type 1 (< -1)", "Type 1 (-1 to -0.25)", "Type 1 (-0.25 to 0)",
  "similar (0-1)", "Type 2 (1-2)", "Type 2 (2-4)", "Type 2 (> 4)"
)

# 3x3 evolution grid of continuous MESS surfaces (list of 9), with the points
# selected so far overlaid, coloured by selection order (dark -> light).
plot_mess_evolution <- function(surfaces, iters, pts, mess_range, title,
                                 pt_cex = 1, pt_pal = \(n) hcl.colors(n, "viridis")) {
  stopifnot(length(surfaces) == 9, length(iters) == 9)
  pt_cols <- pt_pal(nrow(pts))
  layout(layout_mat, heights = c(1, 1, 1, 0.4))
  par(mar = c(1, 1, 2, 0.5), oma = c(0, 0, 3, 0))
  for (k in seq_len(9)) {
    iter <- iters[k]
    plot(surfaces[[k]], main = if (iter == 0) "iter 0" else paste0("iter ", iter),
         col = divergent_pal, range = mess_range, legend = FALSE, axes = FALSE)
    lines(ken)
    if (iter >= 1) {
      psf <- pts[seq_len(iter), , drop = FALSE]
      points(vect(psf, geom = c("longitude", "latitude"), crs = crs(cov_rast_ken_sub)),
             pch = 19, col = pt_cols[seq_len(iter)], cex = pt_cex)
    }
  }
  par(mar = c(2.5, 3, 0.5, 3))
  colorbar_h(mess_range, "MESS (clamped to 1st/99th pctile)")
  mtext(title, outer = TRUE, cex = 1.4, font = 2)
}

# 3x3 evolution grid of discrete-class ExDet surfaces (list of 9), same overlay.
plot_exdet_evolution <- function(surfaces, iters, pts, title,
                                 pt_cex = 1, pt_pal = \(n) hcl.colors(n, "viridis")) {
  stopifnot(length(surfaces) == 9, length(iters) == 9)
  pt_cols <- pt_pal(nrow(pts))
  layout(layout_mat, heights = c(1, 1, 1, 0.5))
  par(mar = c(1, 1, 2, 0.5), oma = c(0, 0, 3, 0))
  for (k in seq_len(9)) {
    iter <- iters[k]
    plot(surfaces[[k]], main = if (iter == 0) "iter 0" else paste0("iter ", iter),
         breaks = exdet_breaks, col = exdet_cols, legend = FALSE, axes = FALSE)
    lines(ken)
    if (iter >= 1) {
      psf <- pts[seq_len(iter), , drop = FALSE]
      points(vect(psf, geom = c("longitude", "latitude"), crs = crs(cov_rast_ken_sub)),
             pch = 19, col = pt_cols[seq_len(iter)], cex = pt_cex)
    }
  }
  par(mar = c(0, 1, 0, 1)); plot.new()
  legend("center", legend = exdet_class_labels, fill = exdet_cols, border = NA,
         ncol = 4, bty = "n", title = "ExDet class", cex = 1.1)
  mtext(title, outer = TRUE, cex = 1.4, font = 2)
}
```

# Data

## Environmental covariates

Ten covariates describe the environment across Kenya; these are the predictors
the selection works in.

```{r}
#| label: fig-covariates
#| fig-width: 12
#| fig-height: 16
par(mfrow = c(5, 2))
for (i in seq_len(nlyr(cov_rast_ken_sub))) {
  plot(cov_rast_ken_sub[[i]], legend = FALSE, axes = FALSE,
       main = names(cov_rast_ken_sub)[i])
}
```

## Existing occurrence records

The reference set: where each species has already been recorded within Kenya.
Coverage is patchy and concentrated, which is exactly why novel environment
remains to be sampled.

```{r}
#| label: fig-points
#| fig-width: 12
#| fig-height: 5
par(mfrow = c(1, 3), mar = c(2, 2, 2, 1))
for (sp in ken_mess_spp) {
  plot(ken); title(main = bquote(italic(.(sp))))
  points(ken_points[ken_points$species == sp, ], pch = 16, col = col_pos, cex = 0.8)
}
```

## Model uncertainty

Per-species coefficient of variation (CV) of the current distribution models.
High CV marks where the models are least constrained — the selection favours
sampling there.

```{r}
#| label: fig-cv
#| fig-width: 12
#| fig-height: 5
cv_range <- range(unlist(lapply(
  ken_mess_spp, \(sp) quantile(values(cv[[sp]]), c(0.01, 0.99), na.rm = TRUE)
)), na.rm = TRUE)
magma_pal <- rev(map.pal("magma", 100))

layout(rbind(1:3, c(4, 4, 4)), heights = c(1, 0.28))
par(mar = c(2, 2, 3, 1))
for (sp in ken_mess_spp) {
  plot(cv[[sp]], col = magma_pal, range = cv_range, legend = FALSE, axes = FALSE)
  title(main = bquote(italic(.(sp)))); lines(ken)
}
par(mar = c(3, 6, 0.5, 6))
colorbar_h(cv_range, "Coefficient of variation (CV)", pal = magma_pal)
```

# Measuring environmental novelty (MESS)

MESS compares each pixel's covariates to the range of values in the reference
set (the occurrence records):

- **Positive (blue)** — similar to the reference; the pixel sits inside the
  observed covariate ranges.
- **Negative (red)** — *novel*; the pixel falls outside the observed range of at
  least one covariate, so a model trained on the reference set is extrapolating.
- **Near zero (white)** — at the edge of the observed environmental space.

The baseline (iteration 0) surfaces below use each species' own records. Note
how similar the three are: the novel (red) regions fall in the same places, so a
**single shared set of survey points** serves all three species.

```{r}
#| label: fig-baseline-mess

mess_baseline <- setNames(
  lapply(ref_vals_0, \(rv) mess(cov_rast_ken_sub, rv)$mess), ken_mess_spp
)
baseline_range <- clamp_range(mess_baseline)

layout(rbind(1:3, c(4, 4, 4)), heights = c(1, 0.28))
par(mar = c(2, 2, 3, 1))
for (sp in ken_mess_spp) {
  plot(mess_baseline[[sp]], col = divergent_pal, range = baseline_range,
       legend = FALSE, axes = FALSE)
  title(main = bquote(italic(.(sp)))); lines(ken)
}
par(mar = c(3, 6, 0.5, 6))
colorbar_h(baseline_range, "MESS (iteration 0)")
```

```{r}
#| label: species-similarity
# Quantify how much the novel (negative) regions coincide across species, as
# justification for a single shared point set.
is_neg   <- lapply(mess_baseline, \(r) values(r)[, 1] < 0)
overlap  <- \(a, b) sum(a & b, na.rm = TRUE) / sum(a | b, na.rm = TRUE)
ov_pairs <- combn(ken_mess_spp, 2, \(p) overlap(is_neg[[p[1]]], is_neg[[p[2]]]))
cat(sprintf(
  "Overlap of novel (negative-MESS) regions between species pairs: %.2f-%.2f\n",
  min(ov_pairs), max(ov_pairs)
))
```

# The selection method

A single shared point is chosen each iteration and added to **every** species'
reference set, so all three evolve around common locations. Two features define
the search:

1. **Weighted by uncertainty.** The selection surface is `MESS × CV`, so the
   pick is the pixel that is both environmentally novel (strongly negative MESS)
   *and* poorly constrained by the current models (high CV). Across species,
   each pixel is driven by its most-dissimilar species.
2. **Distinct cells.** Once a pixel is chosen it is masked out, forcing a *new*
   location every iteration. This matters after novel space is used up:
   otherwise the search re-picks the same near-zero pixel repeatedly, instead of
   spreading points to improve coverage.

Selection runs in two phases, tracked for reporting only (the criterion never
changes): a **novelty** phase while any pixel is still novel (negative MESS
somewhere), then a **coverage** phase once every pixel is inside the envelope.

# Running the selection

The loop below does everything in one pass. At each iteration it holds the
reference set (records + points chosen so far) and, for every species,

- computes **MESS** (drives the pick and the objective trajectory), and
- computes **ExDet** (the independent validation metric, §6),

recording per-iteration totals and keeping every iteration's surface (the rasters
are small). It then selects and adds the next shared point. The evolution grids
below just pick 9 evenly-spaced iterations out of the full history.

```{r}
#| label: run-selection
#| code-fold: show
#| cache: false

n_points   <- 200
grid_iters <- round(seq(0, n_points, length.out = 9))  # 9 iters shown in grids

ref_vals     <- ref_vals_0        # grows as points are added (shared)
chosen_cells <- integer(0)        # grid cells already selected (kept distinct)

selected_points <- data.frame(
  longitude = numeric(0), latitude = numeric(0),
  source_species = character(0), iteration = integer(0), phase = character(0)
)

# Per-iteration scalar trajectories, summed over all cells and species.
traj <- data.frame(
  iteration   = 0:n_points,
  total_mess  = NA_real_,   # sum of MESS (the objective)
  exdet_type1 = NA_real_,   # sum |exdet| where exdet < 0  (univariate novelty)
  exdet_type2 = NA_real_    # sum (exdet - 1) where exdet > 1 (combinational)
)

# Keep every iteration's surface, per species (the rasters are small). Indexed by
# iteration: [[sp]][[i + 1]] is the surface after i points have been added.
mess_surfaces  <- setNames(lapply(ken_mess_spp, \(s) vector("list", n_points + 1)), ken_mess_spp)
exdet_surfaces <- setNames(lapply(ken_mess_spp, \(s) vector("list", n_points + 1)), ken_mess_spp)

novelty_exhausted       <- FALSE
novelty_exhaustion_iter <- NA_integer_

for (i in 0:n_points) {

  # MESS (selection + objective) and ExDet (validation) for every species.
  mess_i  <- setNames(lapply(ref_vals, \(rv) mess(cov_rast_ken_sub, rv)$mess),  ken_mess_spp)
  exdet_i <- setNames(lapply(ref_vals, \(rv) exdet(cov_rast_ken_sub, rv)$exdet), ken_mess_spp)

  mess_vals  <- lapply(mess_i,  \(r) values(r)[, 1])
  exdet_vals <- lapply(exdet_i, \(r) values(r)[, 1])
  traj$total_mess[i + 1]  <- sum(sapply(mess_vals,  \(v) sum(v, na.rm = TRUE)))
  traj$exdet_type1[i + 1] <- sum(sapply(exdet_vals, \(v) sum(abs(v[v < 0]), na.rm = TRUE)))
  traj$exdet_type2[i + 1] <- sum(sapply(exdet_vals, \(v) sum(v[v > 1] - 1,  na.rm = TRUE)))

  for (sp in ken_mess_spp) {
    mess_surfaces[[sp]][[i + 1]]  <- mess_i[[sp]]
    exdet_surfaces[[sp]][[i + 1]] <- exdet_i[[sp]]
  }

  if (i == n_points) break

  # Selection surface: MESS x CV, most-dissimilar species drives each pixel;
  # already-chosen pixels are masked so every pick is distinct.
  sel_cols <- setNames(lapply(ken_mess_spp, \(sp) values(mess_i[[sp]] * cv[[sp]])[, 1]),
                       ken_mess_spp)
  cell_min <- do.call(pmin, c(sel_cols, na.rm = TRUE))
  cell_min[chosen_cells] <- NA
  if (all(is.na(cell_min))) { message("All cells chosen at iter ", i); break }

  sel_cell <- which.min(cell_min)                                   # global pick
  sp_name  <- ken_mess_spp[which.min(sapply(sel_cols, \(v) v[sel_cell]))]

  # Phase bookkeeping: novelty until no negative MESS remains anywhere.
  if (!novelty_exhausted && min(sapply(mess_vals, \(v) min(v, na.rm = TRUE))) >= 0) {
    novelty_exhausted <- TRUE
    novelty_exhaustion_iter <- i
  }

  xy <- xyFromCell(mess_i[[sp_name]], sel_cell)
  selected_points <- rbind(selected_points, data.frame(
    longitude = xy[1, "x"], latitude = xy[1, "y"], source_species = sp_name,
    iteration = i, phase = if (novelty_exhausted) "coverage" else "novelty"
  ))
  chosen_cells <- c(chosen_cells, sel_cell)

  # Add the chosen point to every species' reference set.
  new_vals <- terra::extract(
    cov_rast_ken_sub,
    vect(data.frame(x = xy[1, "x"], y = xy[1, "y"]), geom = c("x", "y"),
         crs = crs(cov_rast_ken_sub)),
    ID = FALSE
  )
  for (sp in ken_mess_spp) ref_vals[[sp]] <- rbind(ref_vals[[sp]], new_vals)
}

cat(sprintf("Selected %d distinct points.\n", nrow(selected_points)))
cat(sprintf("Novelty exhausted at iteration %s.\n",
            ifelse(is.na(novelty_exhaustion_iter), "never", novelty_exhaustion_iter)))
cat(sprintf("Total MESS: %.0f (start) -> %.0f (end).\n",
            traj$total_mess[1], traj$total_mess[nrow(traj)]))
```

# Results

## Novelty is reduced as points are added

The objective — total MESS summed over every pixel and species — rises steeply
as the strongly-novel pixels are absorbed and the covariate envelope widens to
cover them (dashed line: novelty exhausted, no negative MESS left anywhere).
Gains continue past that point: each distinct new pixel keeps shifting the
reference distribution and pulling more area toward higher similarity, so the
full 200-point budget does useful work. The climb is not strictly monotonic —
the criterion picks the *most-dissimilar* pixel, not the one that maximises the
sum — but the trend is clearly upward.

```{r}
#| label: fig-mess-trajectory
#| fig-width: 9
#| fig-height: 5
p <- ggplot(traj, aes(iteration, total_mess)) +
  geom_line(colour = col_pos, linewidth = 0.8) +
  labs(x = "Iteration", y = "Total MESS (summed over cells and species)",
       title = "Total surface MESS as distinct points are added") +
  theme_minimal()
if (!is.na(novelty_exhaustion_iter)) {
  p <- p + geom_vline(xintercept = novelty_exhaustion_iter, linetype = "dashed",
                      colour = col_neg) +
    annotate("text", x = novelty_exhaustion_iter,
             y = min(traj$total_mess, na.rm = TRUE), label = "novelty exhausted",
             hjust = -0.05, vjust = 0, colour = col_neg, size = 3)
}
p
```

## The 200 selected points

The deliverable: 200 shared survey locations. Points chosen during the
**novelty** phase (red) target environmentally extrapolated areas; **coverage**
points (blue) spread through the remaining space once novelty is used up.

```{r}
#| label: fig-selected-points
#| fig-width: 7
#| fig-height: 8
plot(ken, main = sprintf("%d selected survey points", nrow(selected_points)))
phase_cols <- c(novelty = col_neg, coverage = col_pos)
points(vect(selected_points, geom = c("longitude", "latitude"), crs = crs(cov_rast_ken_sub)),
       pch = 19, cex = 0.9, col = phase_cols[selected_points$phase])
legend("bottomleft", legend = names(phase_cols), pch = 19, col = phase_cols,
       bty = "n", title = "phase", inset = 0.02)
```

```{r}
#| label: points-by-phase
selected_points |> count(phase, name = "points") |> knitr::kable()
```

## Evolution of the novelty surface

How the MESS surface fills in for the representative species
(*`r rep_sp`*; the others are near-identical). Iteration 0, iteration 200, and 7
evenly-spaced steps between, with the points selected so far overlaid (dark =
first selected, yellow = last).

```{r}
#| label: fig-mess-evolution
#| fig-width: 10
#| fig-height: 11
mess_shown <- mess_surfaces[[rep_sp]][grid_iters + 1]   # 9 surfaces to display
plot_mess_evolution(
  surfaces   = mess_shown,
  iters      = grid_iters,
  pts        = selected_points,
  mess_range = clamp_range(mess_shown),
  title      = bquote(italic(.(rep_sp)) * " — MESS over the selected points")
)
```

## Exported points

The full point set with priority (selection order), source species, and phase,
written to `k_m_r_selected_points.csv`.

```{r}
#| label: export-points
selected_points <- selected_points |>
  mutate(priority = row_number(), .before = 1)
write_csv(selected_points, "k_m_r_selected_points.csv")
head(selected_points, 10) |> knitr::kable(digits = 3)
```

# Validation with ExDet

MESS drives the selection, so it is not an independent check. **ExDet**
(Extrapolation Detection) is a second, richer novelty metric computed on the
same growing reference set (already recorded in the loop above). It splits
novelty into two kinds, encoded in a single value per pixel:

- **`exdet < 0` — Type 1 (univariate):** outside the observed range of at least
  one covariate. This is the same regime MESS flags as negative.
- **`0 ≤ exdet ≤ 1` — similar:** in range univariately and inside the
  combinational (Mahalanobis) envelope.
- **`exdet > 1` — Type 2 (combinational):** every covariate is individually in
  range, but their *combination* is novel.

## Novelty trajectories

**Type 1** novelty collapses to zero within the first ~16 points (grey dotted
line): each extreme point widens some covariate's observed range, quickly
bringing every pixel univariately in-range — exactly what the MESS objective is
driving.

**Type 2** novelty only wobbles *while Type 1 is still collapsing* (all the
movement is left of the grey line), then settles into a slow decline. It never
reaches zero: combinational novelty is much harder to eliminate than univariate
novelty, so some pixels remain novel *combinations* even after 200 points.

::: {.callout-note collapse="true"}
## Why Type 2 rises before the grey line (an ExDet artefact, not created novelty)

A pixel reports its ExDet score as `ifelse(nt1 == 0, nt2, nt1)`, so while it is
univariately novel its combinational novelty is **masked**; as Type 1 collapses,
those pixels reveal their Type 2 values and swell the sum. On top of that,
`nt2 = mahalanobis / mah_max` is normalised by the *most-extreme reference
point*, so adding an extreme point transiently inflates `mah_max` and rescales
the whole surface. Both effects are active only while the reference set is
univariately unsaturated — i.e. left of the grey line — after which Type 2
declines cleanly.
:::

```{r}
#| label: fig-exdet-trajectory
#| fig-width: 9
#| fig-height: 5
exdet_long <- traj |>
  select(iteration, exdet_type1, exdet_type2) |>
  pivot_longer(-iteration, names_to = "novelty", values_to = "total") |>
  mutate(novelty = recode(novelty, exdet_type1 = "Type 1 (univariate)",
                          exdet_type2 = "Type 2 (combinational)"))
type1_zero_iter <- traj$iteration[which(traj$exdet_type1 == 0)[1]]

p <- ggplot(exdet_long, aes(iteration, total, colour = novelty)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c("Type 1 (univariate)" = "#D6604D",
                                 "Type 2 (combinational)" = "#762A83")) +
  labs(x = "Iteration", y = "Total novelty (summed over cells and species)",
       colour = NULL, title = "ExDet novelty as points are added") +
  theme_minimal() + theme(legend.position = "top")
if (!is.na(type1_zero_iter)) {
  p <- p + geom_vline(xintercept = type1_zero_iter, linetype = "dotted",
                      colour = "grey50", linewidth = 0.6)
}
p
```

## Surface evolution

The ExDet surface for the representative species. Type 1 (red) extrapolation
covers most of Kenya at iteration 0 and is essentially gone within the first
grid step; a residual Type 2 (purple) combinational-novelty patch in the
highlands persists and slowly shrinks.

```{r}
#| label: fig-exdet-evolution
#| fig-width: 10
#| fig-height: 11
plot_exdet_evolution(
  surfaces = exdet_surfaces[[rep_sp]][grid_iters + 1],   # 9 surfaces to display
  iters    = grid_iters,
  pts      = selected_points,
  title    = bquote(italic(.(rep_sp)) * " — ExDet over the selected points")
)
```
