---
title: "Selecting survey points for *Anopheles* in Kenya"
subtitle: "Three-phase sampling for environmental novelty weighted by uncertainty"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "code"
    fig-align: center
    theme: lux
execute:
  warning: false
  message: false
---

# Summary

Vector Atlas records do not cover Kenya's environments evenly. Records for three
malaria vectors — *An. arabiensis*, *An. gambiae* and *An. funestus* — sit mostly
in the wetter, farmed south-west, and the arid north and east barely appear at
all. Where a model has never seen an environment it has to extrapolate, and those
are exactly the places its predictions are least trustworthy. This analysis
chooses 50 places to survey next, to close that gap, weighted by where the models
are least certain. Some of these sites may be in inaccessible or insecure
locations. Skip them and move on.

**Measuring the gap.** For every 5x5 km pixel in Kenya we ask how unlike the
places we have already surveyed its environment is. We use three measures of
environmental novelty: the Multivariate Environmental Similarity Surface (MESS,
Elith et al. 2010), ExDet (Mesgaran et al. 2014), and Shape (Velazco et al.
2024). MESS and ExDet Type 1 find pixels that fall outside the range of anything
we have sampled so far, e.g. more shrubland or less tree cover than at any
existing record. ExDet Type 2 finds pixels that are within the existing range of
every variable individually, but have novel combinations of those variables.
Shape sets that distinction aside and reports one number: how far a pixel sits,
in environmental terms, from the nearest place we have already surveyed.

**Choosing a site.** We multiply each measure by how uncertain the current models
are at that pixel via the coefficient of variation, so the highest-weighted sites
are both environmentally unfamiliar and poorly predicted. To select a new survey
point, we take the highest-scoring pixel, add it to the set of existing points,
and recalculate all three measures before making the next choice. This repeats
until we have 50 points. No pixel appears twice, and a single shared set of
points serves all three species; at each step the site with the highest weight
value across the three species is the one chosen.

**Three phases.** Our selection process works through types of novelty in the
order it can remove them: first the out-of-range pixels, then the unusual
combinations, then whatever remains. The last phase works differently. We first
divide Kenya into 150 multivariate environmental clusters, each with a single
typical point (its "medoid"). The design picks among those, ranking them by
model uncertainty and Shape value.

**Result.** A ranked list of 50 sites (`k_m_r_selected_points.csv`), maps
tracking how much of Kenya's environment is still unrepresented as we add points,
a record of which species drove each choice, and a map of which environment types
remain without any sampling at all.

```{r}
#| label: setup

library(targets)
library(terra)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(bssdm)
library(flexsdm)

# ---- Spatial inputs (targets store) ----------------------------------------
tar_load(
  c(covariate_rast_5, model_data_spatial, ken, target_covariate_names),
  store = "~/Documents/tki_work/vector_atlas/va_multispecies_sdm/_targets/"
)

# Covariate stack cropped/masked to Kenya, reduced to the modelling covariates.
cov_rast_ken_sub <- crop(covariate_rast_5, ken) |>
  mask(ken) |>
  subset(target_covariate_names)

ken_mess_spp <- c("arabiensis", "gambiae", "funestus")

# ---- Design parameters ------------------------------------------------------
# Defined here so every mention of them downstream can be read off the values
# rather than typed in (the numbers in the summary above are the exception -
# inline code cannot run before this chunk).
n_points <- 50     # survey locations to select
n_clust  <- 150    # environmental clusters for the phase-3 medoid pool

# ---- Model uncertainty: per-species coefficient of variation ----------------
cv <- rast(paste0(
  "~/Documents/tki_work/vector_atlas/va_multispecies_sdm/",
  "spartan_model_comparison/m6/m6_mask_p_cv.tif"
)) |>
  crop(ken) |>
  mask(ken)
cv <- cv[[ken_mess_spp]]
cv <- resample(cv, cov_rast_ken_sub, method = "bilinear")   # align to covariates

# ---- Existing occurrence points within Kenya, per species -------------------
ken_points <- vect(
  model_data_spatial |>
    filter(!inferred) |> 
    select(species, longitude, latitude) |>
    filter(species %in% ken_mess_spp),
  geom = c("longitude", "latitude"), crs = crs(ken)
)
ken_points <- ken_points[ken, ]

# Reference environmental values per species (their observation points only).
ref_vals_0 <- setNames(
  lapply(ken_mess_spp, \(sp) {
    terra::extract(cov_rast_ken_sub, ken_points[ken_points$species == sp, ],
                   ID = FALSE) |> na.omit()
  }),
  ken_mess_spp
)

# Per-species CV as plain vectors over the grid, reused as the selection weight.
cv_cols <- setNames(lapply(ken_mess_spp, \(sp) values(cv[[sp]])[, 1]), ken_mess_spp)

# Land cells (complete covariates): the pool the k-medoid partition works over.
land_cells <- which(stats::complete.cases(values(cov_rast_ken_sub)))
```

```{r}
#| label: helpers

# Palette / colours reused throughout. Red = novel, blue = similar.
col_neg <- "#D6604D"
col_pos <- "#4393C3"
divergent_pal <- colorRampPalette(c(col_neg, "white", col_pos))(255)
magma_pal     <- map.pal("magma", 100)        # dark = in-niche, bright = novel

# One colour per selection phase, reused in maps and trajectory markers.
phase_cols <- c(type1 = "#D6604D", type2 = "#762A83", medoid = "#F98C0A")
phase_labels <- c(type1  = "Phase 1: Type 1 (univariate)",
                  type2  = "Phase 2: Type 2 (combinational)",
                  medoid = "Phase 3: k-medoid clusters (CV x Shape)")

# 3x3 map grid plus a colourbar/legend strip below it.
layout_mat <- rbind(matrix(1:9, nrow = 3, byrow = TRUE), rep(10, 3))

# Symmetric colour range clamped to the 1st/99th percentile of a list of
# rasters, so outliers don't wash out the variation.
clamp_range <- function(rasters) {
  absmax <- max(abs(unlist(lapply(
    rasters, \(r) quantile(values(r), c(0.01, 0.99), na.rm = TRUE)
  ))))
  c(-absmax, absmax)
}

# Horizontal colourbar spanning a panel.
colorbar_h <- function(rng, label, pal = divergent_pal) {
  vals <- seq(rng[1], rng[2], length.out = length(pal))
  image(x = vals, y = 1, z = matrix(vals, ncol = 1), col = pal,
        axes = FALSE, xlab = label, ylab = "")
  axis(1)
}

# One row of three species maps (continuous surface) over a shared colourbar.
# terra's plot() rejects bquote() in `main`, so titles go on afterwards with
# title() - see the same pattern in the CV and occurrence figures.
plot_species_row <- function(rl, rng, label, pal = divergent_pal) {
  layout(rbind(1:3, c(4, 4, 4)), heights = c(1, 0.28))
  par(mar = c(2, 2, 3, 1))
  for (sp in ken_mess_spp) {
    plot(rl[[sp]], col = pal, range = rng, legend = FALSE, axes = FALSE)
    title(main = bquote(italic(.(sp)))); lines(ken)
  }
  par(mar = c(3, 6, 0.5, 6))
  colorbar_h(rng, label, pal = pal)
}

# ---- Novelty metrics --------------------------------------------------------

# Shape: multivariate extrapolation surface for a reference set, via
# flexsdm::extra_eval (the "Shape" metric of Velazco et al. 2024). Presence-only
# reference (all rows pr_ab = 1), so extrapolation is measured against that
# environmental cloud alone. What it returns, precisely: the Mahalanobis distance
# from a cell to the NEAREST reference record (not to the cloud's centroid),
# divided by the mean Mahalanobis distance of the reference records from their
# own centroid, x 100. So it is a nearest-neighbour distance expressed as a
# percentage of the reference set's own spread: 100 means "as far from the
# closest record as a typical record is from the middle of the record cloud".
# Always >= 0, ~0 at a record, rising with distance from the nearest one.
shape_eval <- function(rv) {
  td <- as.data.frame(rv)
  td$pr_ab <- 1L
  flexsdm::extra_eval(
    training_data   = td,
    pr_ab           = "pr_ab",
    projection_data = cov_rast_ken_sub,
    metric          = "mahalanobis"
  )
}

# ExDet novelty magnitudes (positive = novel; NA where a cell is not that type,
# so a phase is "exhausted" once no eligible cell remains).
type1_mag <- function(v) ifelse(v < 0, -v,    NA_real_)   # univariate:    exdet < 0
type2_mag <- function(v) ifelse(v > 1, v - 1, NA_real_)   # combinational: exdet > 1

# Cross-species reducer: a shared point => the most-extreme species drives a cell.
pmax_na <- function(cols) suppressWarnings(do.call(pmax, c(cols, na.rm = TRUE)))

# Wrap a vector of cell values back into a raster for plotting.
as_rast <- function(v) {
  r <- rast(cov_rast_ken_sub[[1]]); values(r) <- v; r
}

# ExDet is shown with deterministic discrete classes (fixed `breaks`), NOT a
# continuous palette: terra re-interpolates a long `col` vector and bleeds the
# near-zero colour into the most *typical* cells, which multiply as points are
# added and look misleadingly like growing novelty. Fixed breaks keep the
# [0, 1] "similar" band grey. Lower "similar" bound is -1e-9 so exdet == 0 is
# grey, not the reddest class.
exdet_breaks <- c(-Inf, -1, -0.25, -1e-9, 1, 2, 4, Inf)
exdet_cols   <- c("#67001F", "#D6604D", "#FDDBC7",   # Type 1: strong -> weak
                  "#f0f0f0",                          # similar
                  "#E7D4E8", "#9970AB", "#40004B")    # Type 2: weak -> strong
exdet_class_labels <- c(
  "Type 1 (< -1)", "Type 1 (-1 to -0.25)", "Type 1 (-0.25 to 0)",
  "similar (0-1)", "Type 2 (1-2)", "Type 2 (2-4)", "Type 2 (> 4)"
)

# One row of three species ExDet maps over a shared discrete-class legend.
plot_exdet_row <- function(rl) {
  layout(rbind(1:3, c(4, 4, 4)), heights = c(1, 0.32))
  par(mar = c(2, 2, 3, 1))
  for (sp in ken_mess_spp) {
    plot(rl[[sp]], breaks = exdet_breaks, col = exdet_cols,
         legend = FALSE, axes = FALSE)
    title(main = bquote(italic(.(sp)))); lines(ken)
  }
  par(mar = c(0, 1, 0, 1)); plot.new()
  legend("center", legend = exdet_class_labels, fill = exdet_cols, border = NA,
         ncol = 4, bty = "n", title = "ExDet class", cex = 1.1)
}

# ---- k-medoid partition -----------------------------------------------------

# k-medoids by the alternating (Voronoi) algorithm, seeded from k-means:
# assign every cell to its nearest medoid, then replace each medoid by the
# cluster member with the smallest total distance to the rest of its cluster.
# Unlike a k-means centroid, a medoid is always a REAL land cell, so it can be
# visited; and it is the most *typical* member of its cluster rather than an
# extreme one. Clusters larger than `pool` are costed against a random subset of
# `pool` members to keep the update linear in the number of cells.
kmedoids <- function(Z, k, seed = 1, iter_max = 15, pool = 400) {
  set.seed(seed)
  n  <- nrow(Z)
  km <- stats::kmeans(Z, centers = k, iter.max = 50, nstart = 5)

  # Start from the cell nearest each k-means centroid.
  med <- vapply(seq_len(k), function(cl) {
    idx <- which(km$cluster == cl)
    ctr <- matrix(km$centers[cl, ], length(idx), ncol(Z), byrow = TRUE)
    idx[which.min(rowSums((Z[idx, , drop = FALSE] - ctr)^2))]
  }, integer(1))

  cl_of <- km$cluster
  for (it in seq_len(iter_max)) {
    d2  <- sapply(med, \(r) rowSums((Z - rep(Z[r, ], each = n))^2))
    new_cl <- max.col(-d2, ties.method = "first")

    med_new <- vapply(seq_len(k), function(cl) {
      idx <- which(new_cl == cl)
      if (!length(idx)) return(med[cl])
      ref  <- if (length(idx) > pool) sample(idx, pool) else idx
      cost <- sapply(idx, \(i) sum(sqrt(rowSums(
        (Z[ref, , drop = FALSE] - rep(Z[i, ], each = length(ref)))^2))))
      idx[which.min(cost)]
    }, integer(1))

    changed <- sum(med_new != med)
    med <- med_new; cl_of <- new_cl
    if (changed == 0) break
  }
  list(k = k, medoid = med, cluster = cl_of, iters = it)
}

# ---- Evolution grids --------------------------------------------------------
# Each takes 9 surfaces + the iteration numbers they correspond to, and overlays
# the points selected so far, coloured by selection order (dark -> light).

# 3x3 grid of continuous MESS surfaces (divergent palette).
plot_mess_evolution <- function(surfaces, iters, pts, mess_range, title,
                                 pt_cex = 1, pt_pal = \(n) hcl.colors(n, "viridis")) {
  stopifnot(length(surfaces) == 9, length(iters) == 9)
  pt_cols <- pt_pal(nrow(pts))
  layout(layout_mat, heights = c(1, 1, 1, 0.4))
  par(mar = c(1, 1, 2, 0.5), oma = c(0, 0, 3, 0))
  for (k in seq_len(9)) {
    iter <- iters[k]
    plot(surfaces[[k]], main = if (iter == 0) "iter 0" else paste0("iter ", iter),
         col = divergent_pal, range = mess_range, legend = FALSE, axes = FALSE)
    lines(ken)
    if (iter >= 1) {
      psf <- pts[seq_len(iter), , drop = FALSE]
      points(vect(psf, geom = c("longitude", "latitude"), crs = crs(cov_rast_ken_sub)),
             pch = 19, col = pt_cols[seq_len(iter)], cex = pt_cex)
    }
  }
  par(mar = c(2.5, 3, 0.5, 3))
  colorbar_h(mess_range, "MESS (clamped to 1st/99th pctile)")
  mtext(title, outer = TRUE, cex = 1.4, font = 2)
}

# 3x3 grid of discrete-class ExDet surfaces.
plot_exdet_evolution <- function(surfaces, iters, pts, title,
                                 pt_cex = 1, pt_pal = \(n) hcl.colors(n, "viridis")) {
  stopifnot(length(surfaces) == 9, length(iters) == 9)
  pt_cols <- pt_pal(nrow(pts))
  layout(layout_mat, heights = c(1, 1, 1, 0.5))
  par(mar = c(1, 1, 2, 0.5), oma = c(0, 0, 3, 0))
  for (k in seq_len(9)) {
    iter <- iters[k]
    plot(surfaces[[k]], main = if (iter == 0) "iter 0" else paste0("iter ", iter),
         breaks = exdet_breaks, col = exdet_cols, legend = FALSE, axes = FALSE)
    lines(ken)
    if (iter >= 1) {
      psf <- pts[seq_len(iter), , drop = FALSE]
      points(vect(psf, geom = c("longitude", "latitude"), crs = crs(cov_rast_ken_sub)),
             pch = 19, col = pt_cols[seq_len(iter)], cex = pt_cex)
    }
  }
  par(mar = c(0, 1, 0, 1)); plot.new()
  legend("center", legend = exdet_class_labels, fill = exdet_cols, border = NA,
         ncol = 4, bty = "n", title = "ExDet class", cex = 1.1)
  mtext(title, outer = TRUE, cex = 1.4, font = 2)
}

# 3x3 grid of continuous Shape surfaces (sequential magma; bright = more novel).
plot_shape_evolution <- function(surfaces, iters, pts, shape_range, title,
                                 pt_cex = 1, pt_pal = \(n) hcl.colors(n, "viridis")) {
  stopifnot(length(surfaces) == 9, length(iters) == 9)
  pt_cols <- pt_pal(nrow(pts))
  layout(layout_mat, heights = c(1, 1, 1, 0.4))
  par(mar = c(1, 1, 2, 0.5), oma = c(0, 0, 3, 0))
  for (k in seq_len(9)) {
    iter <- iters[k]
    plot(surfaces[[k]], main = if (iter == 0) "iter 0" else paste0("iter ", iter),
         col = magma_pal, range = shape_range, legend = FALSE, axes = FALSE)
    lines(ken)
    if (iter >= 1) {
      psf <- pts[seq_len(iter), , drop = FALSE]
      points(vect(psf, geom = c("longitude", "latitude"), crs = crs(cov_rast_ken_sub)),
             pch = 19, col = pt_cols[seq_len(iter)], cex = pt_cex)
    }
  }
  par(mar = c(2.5, 3, 0.5, 3))
  colorbar_h(shape_range, "Shape extrapolation (higher = more novel)", pal = magma_pal)
  mtext(title, outer = TRUE, cex = 1.4, font = 2)
}
```

# Data

## Environmental covariates

`r nlyr(cov_rast_ken_sub)` covariates describe the environment across Kenya;
these are the predictors the selection works in.

```{r}
#| label: fig-covariates
#| fig-width: 12
#| fig-height: 16
par(mfrow = c(5, 2))
for (i in seq_len(nlyr(cov_rast_ken_sub))) {
  plot(cov_rast_ken_sub[[i]], legend = FALSE, axes = FALSE,
       main = names(cov_rast_ken_sub)[i])
}
```

## Existing occurrence records

The reference set: where we already have records for each species within Kenya.
Coverage is patchy and concentrated, which is exactly why so much novel
environment still remains.

```{r}
#| label: fig-points
#| fig-width: 12
#| fig-height: 5
par(mfrow = c(1, 3), mar = c(2, 2, 2, 1))
for (sp in ken_mess_spp) {
  plot(ken); title(main = bquote(italic(.(sp))))
  points(ken_points[ken_points$species == sp, ], pch = 16, col = col_pos, cex = 0.8)
}
```

## Model uncertainty

Per-species coefficient of variation (CV) of the current distribution models.
High CV marks where the models have least to go on — the selection favours
sampling there.

```{r}
#| label: fig-cv
#| fig-width: 12
#| fig-height: 5
cv_range <- range(unlist(lapply(
  ken_mess_spp, \(sp) quantile(values(cv[[sp]]), c(0.01, 0.99), na.rm = TRUE)
)), na.rm = TRUE)
cv_pal <- rev(map.pal("magma", 100))

layout(rbind(1:3, c(4, 4, 4)), heights = c(1, 0.28))
par(mar = c(2, 2, 3, 1))
for (sp in ken_mess_spp) {
  plot(cv[[sp]], col = cv_pal, range = cv_range, legend = FALSE, axes = FALSE)
  title(main = bquote(italic(.(sp)))); lines(ken)
}
par(mar = c(3, 6, 0.5, 6))
colorbar_h(cv_range, "Coefficient of variation (CV)", pal = cv_pal)
```

# Environmental novelty at the existing records

Before choosing any point, we ask the same question of every pixel in Kenya:
*how unlike the environments we have already recorded this species in is it?*
The three metrics below answer it in three different ways, each against that
species' own reference records.

```{r}
#| label: baseline-metrics

# All three metrics at iteration 0, for each species, against its own records.
mess_out       <- setNames(lapply(ref_vals_0, \(rv) mess(cov_rast_ken_sub, rv)), ken_mess_spp)
mess_baseline  <- lapply(mess_out, \(m) m$mess)
mod_baseline   <- lapply(mess_out, \(m) m$mod)
exdet_baseline <- setNames(lapply(ref_vals_0, \(rv) exdet(cov_rast_ken_sub, rv)$exdet), ken_mess_spp)
shape_baseline <- setNames(lapply(ref_vals_0, \(rv) shape_eval(rv)), ken_mess_spp)

n_land <- length(land_cells)
pct    <- \(x) 100 * sum(x, na.rm = TRUE) / n_land
```

## MESS — outside the *range* of the records

**In plain language:** for each covariate, MESS asks whether a pixel's value
falls inside the minimum–maximum range of the records, and reports the single
worst covariate as a percentage of that range. Positive (blue) means every
covariate falls inside the range the records span — the pixel is
*interpolation*. Negative (red) means at least one covariate lies outside
anything we have sampled, and the more negative the further outside. MESS is fast
and easy to explain, but it only ever looks at one covariate at a time.

```{r}
#| label: fig-mess-baseline
#| fig-width: 12
#| fig-height: 5
baseline_range <- clamp_range(mess_baseline)
plot_species_row(mess_baseline, baseline_range, "MESS (negative = outside sampled range)")
```

Which covariate is responsible? The maps below give the *most dissimilar
variable* (MOD) at every cell where MESS is negative — the covariate whose value
sits furthest outside the range of that species' records. Cells that are inside
the sampled range on every covariate have no MOD and stay blank.

```{r}
#| label: fig-mod
#| fig-width: 12
#| fig-height: 5

# Fixed category order and palette, identical across the three panels, so a
# colour always means the same covariate rather than "whatever ranked here".
# Seven covariates account for essentially all the novelty; the rest fold into
# "other". Hues were chosen to clear all-pairs colour-vision separation against
# the pale "not novel" background (a map puts any two categories side by side,
# so every pair has to be separable, not just neighbouring legend entries).
mod_levels <- c("shrubs", "trees", "grassland", "cropland",
                "built", "footprint", "prox_to_sea")
mod_cols   <- c("#884422", "#228866", "#aa8800", "#ee4488",
                "#880088", "#4444cc", "#0088cc")
mod_other_col <- "#7a7a76"
mod_bg_col    <- "#f2f2f0"
mod_fill      <- c(mod_cols, mod_other_col)

# Per-species tally over the novel cells; kept for the numbers quoted in the
# text below, and for deciding what "other" has to absorb.
mod_counts <- setNames(lapply(ken_mess_spp, \(sp) {
  novel <- values(mess_baseline[[sp]])[, 1] < 0
  modv  <- values(mod_baseline[[sp]])[, 1][which(novel)]
  sort(table(names(cov_rast_ken_sub)[modv]), decreasing = TRUE)
}), ken_mess_spp)

mod_tab <- do.call(rbind, lapply(ken_mess_spp, \(sp) {
  tt <- mod_counts[[sp]]
  data.frame(
    species        = sp,
    `novel cells (%)` = round(pct(values(mess_baseline[[sp]])[, 1] < 0), 1),
    `worst covariate` = names(tt)[1],
    `share of novel cells (%)` = round(100 * tt[[1]] / sum(tt), 1),
    `next covariate` = names(tt)[2],
    check.names = FALSE
  )
}))

# What "other" absorbs, and the largest share it ever reaches.
mod_other_names <- sort(unique(unlist(lapply(
  mod_counts, \(tt) setdiff(names(tt), mod_levels)))))
mod_other_max <- max(sapply(mod_counts, \(tt) {
  o <- tt[!names(tt) %in% mod_levels]
  if (!length(o)) 0 else 100 * max(o) / sum(tt)
}))

# Land background: every cell with complete covariates, novel or not.
land_mask <- rast(cov_rast_ken_sub[[1]])
values(land_mask) <- NA_real_
values(land_mask)[land_cells] <- 1

# MOD as a category code, NA wherever the cell is not univariately novel.
mod_code_rast <- function(sp) {
  novel <- values(mess_baseline[[sp]])[, 1] < 0
  nm    <- names(cov_rast_ken_sub)[values(mod_baseline[[sp]])[, 1]]
  code  <- match(nm, mod_levels)
  code[is.na(code) & !is.na(nm)] <- length(mod_levels) + 1L   # "other"
  code[is.na(novel) | !novel] <- NA
  r <- rast(cov_rast_ken_sub[[1]]); values(r) <- code; r
}
mod_rasts <- setNames(lapply(ken_mess_spp, mod_code_rast), ken_mess_spp)

layout(rbind(1:3, c(4, 4, 4)), heights = c(1, 0.22))
par(mar = c(0.5, 1, 2.5, 1))
for (sp in ken_mess_spp) {
  plot(land_mask, col = mod_bg_col, legend = FALSE, axes = FALSE)
  # Fixed breaks, one per category, so a code always draws the same colour.
  plot(mod_rasts[[sp]], breaks = seq(0.5, length(mod_fill) + 0.5, 1),
       col = mod_fill, legend = FALSE, axes = FALSE, add = TRUE)
  title(main = bquote(italic(.(sp)))); lines(ken)
}
par(mar = c(0, 1, 0, 1)); plot.new()
legend("center", legend = c(mod_levels, "other", "not novel"),
       fill = c(mod_fill, mod_bg_col),
       border = c(rep(NA, length(mod_fill)), "grey60"),
       ncol = 5, bty = "n", cex = 1.1,
       title = "most dissimilar variable at cells where MESS < 0")
```

The covariates folded into "other"
(`r paste(mod_other_names, collapse = ", ")`) are the most dissimilar variable
for at most `r round(mod_other_max, 1)`% of novel cells in any species.

## ExDet — range novelty *and* novel combinations

**In plain language:** ExDet keeps the MESS idea (Type 1, `exdet < 0`: at least
one covariate outside the range the records cover) and adds the part MESS is
blind to (Type 2, `exdet > 1`: every covariate sits individually in range, but no
record shows the *combination* — high tree cover together with high human
footprint would be one such combination, though this report does not check which
combinations are responsible here). Values between 0 and 1 mean "similar" — the
pixel sits inside the environmental cloud the records span. The map uses fixed
classes, reds for Type 1, purples for Type 2, grey for similar.

A reporting rule matters for everything that follows: a pixel's ExDet score is
`nt1` where the pixel is Type 1 and `nt2` only where it is not. Type 2 is
therefore invisible wherever Type 1 applies, which is most of Kenya at the
outset — so the small Type-2 area below is a floor, not a measure of how much
combinational novelty exists.

```{r}
#| label: fig-exdet-baseline
#| fig-width: 12
#| fig-height: 5.5
plot_exdet_row(exdet_baseline)
```

## Shape — one continuous distance from the nearest record

**In plain language:** Shape drops the Type 1 / Type 2 distinction and reports a
single number per pixel: how far that pixel is, in environmental terms, from the
nearest existing record. Distance is Mahalanobis, so it accounts for correlations
between covariates, and Shape expresses it as a percentage of the reference set's
own spread — the mean distance of the records from the middle of their own cloud.
A value of 100 means "as far from the closest record as a typical record is from
the centre of the record cloud"; the values of several thousand below therefore
mean tens of times the typical spread.

Two consequences are worth holding on to. It is a *nearest-neighbour* distance,
so a single new record inside a novel region collapses Shape across that whole
region — which is exactly what happens at iteration 20 below. And it is
always `≥ 0` and defined everywhere, so unlike the ExDet types it never "runs
out"; that is why it is the measure phase 3 keeps shrinking.

```{r}
#| label: fig-shape-baseline
#| fig-width: 12
#| fig-height: 5
shape_hi <- max(unlist(lapply(shape_baseline, \(r) quantile(values(r), 0.99, na.rm = TRUE))))
plot_species_row(shape_baseline, c(0, shape_hi),
                 "Shape (Mahalanobis extrapolation; higher = more novel)",
                 pal = magma_pal)
```

## What is outside the reference values, by species

```{r}
#| label: tbl-novelty-baseline
novelty_tab <- do.call(rbind, lapply(ken_mess_spp, \(sp) {
  mv <- values(mess_baseline[[sp]])[, 1]
  ev <- values(exdet_baseline[[sp]])[, 1]
  sv <- values(shape_baseline[[sp]])[, 1]
  data.frame(
    species = sp,
    `records` = nrow(ref_vals_0[[sp]]),
    `MESS < 0 (%)`      = round(pct(mv < 0), 1),
    `ExDet Type 1 (%)`  = round(pct(ev < 0), 1),
    `ExDet Type 2 (%)`  = round(pct(ev > 1), 1),
    `similar (%)`       = round(pct(ev >= 0 & ev <= 1), 1),
    `median Shape`      = round(median(sv, na.rm = TRUE), 2),
    `max Shape`         = round(max(sv, na.rm = TRUE), 1),
    check.names = FALSE
  )
}))
knitr::kable(novelty_tab, row.names = FALSE)

# Ranges quoted in the text below, read off the tables so they cannot drift.
rng_txt  <- \(x, fmt = "%.1f") sprintf(paste0(fmt, "-", fmt), min(x), max(x))
t1_txt   <- rng_txt(novelty_tab[["ExDet Type 1 (%)"]], "%.0f")
t2_txt   <- rng_txt(novelty_tab[["ExDet Type 2 (%)"]], "%.2f")
mod1_txt <- rng_txt(mod_tab[["share of novel cells (%)"]], "%.0f")
t2_worst <- novelty_tab$species[which.max(novelty_tab[["ExDet Type 2 (%)"]])]
```

The three species tell nearly the same story. Univariate (Type 1) novelty is
the dominant problem: `r t1_txt`% of Kenya lies outside the sampled range of at
least one covariate — everything but the well-surveyed south-west. For all three
species the covariate most often responsible is *shrubs* (`r mod1_txt`% of novel
pixels; the runner-up differs by species, see the maps above): the records
simply do not span the arid shrubland that covers most of the country.

Combinational (Type 2) novelty looks rare at this stage — `r t2_txt`% of
pixels, at most a few dozen cells, largest for *`r t2_worst`* — and sits mostly
in the north-west around Lake Turkana, with a scatter further south. That
near-absence is largely the masking rule noted above: ExDet only reports Type 2
where a pixel is not already Type 1, and Type 1 covers most of the country. Once
phase 1 strips the range novelty away, Type 2 grows by more than an order of
magnitude before it collapses — see the trajectories.

Shape rises away from the south-west, peaking in the same north-western block.
Because the novel regions coincide closely across species, a single shared set
of survey points serves all three.

```{r}
#| label: species-similarity
# Quantify how much the novel (negative) regions coincide across species.
is_neg   <- lapply(mess_baseline, \(r) values(r)[, 1] < 0)
overlap  <- \(a, b) sum(a & b, na.rm = TRUE) / sum(a | b, na.rm = TRUE)
ov_pairs <- combn(ken_mess_spp, 2, \(p) overlap(is_neg[[p[1]]], is_neg[[p[2]]]))
cat(sprintf(
  "Overlap of novel (negative-MESS) regions between species pairs: %.2f-%.2f\n",
  min(ov_pairs), max(ov_pairs)
))
```

# Weighting novelty by uncertainty

Novelty alone would send researchers to the most extreme corner of the covariate
space regardless of whether the models need help there. Every phase therefore
multiplies its novelty measure by the per-pixel CV, so a pixel is attractive
only if it is *both* environmentally new *and* one the current models predict
poorly. The three products below are exactly the surfaces the three phases
maximise (blank = the pixel is not of that novelty type, so it never becomes a
candidate for that phase).

```{r}
#| label: baseline-weights
exdet_vals_0 <- lapply(exdet_baseline, \(r) values(r)[, 1])
shape_vals_0 <- lapply(shape_baseline, \(r) values(r)[, 1])

w_t1    <- setNames(lapply(ken_mess_spp, \(sp) cv_cols[[sp]] * type1_mag(exdet_vals_0[[sp]])), ken_mess_spp)
w_t2    <- setNames(lapply(ken_mess_spp, \(sp) cv_cols[[sp]] * type2_mag(exdet_vals_0[[sp]])), ken_mess_spp)
w_shape <- setNames(lapply(ken_mess_spp, \(sp) cv_cols[[sp]] * shape_vals_0[[sp]]),            ken_mess_spp)

w_rng <- \(wl) c(0, max(unlist(lapply(wl, \(v) quantile(v, 0.99, na.rm = TRUE))), na.rm = TRUE))
```

## CV × ExDet Type 1 — the phase 1 weighting

Bright cells lie far outside the sampled range *and* the models predict them
poorly. The Type-1 region covers most of the country outside the south-west, but
the CV weight concentrates the attractive part of it in the north and east. Note
how differently the three species score the *same* pixels: because the
cross-species reducer is `pmax`, whichever species is most
extrapolated-and-uncertain at a pixel sets its score, and we record it as having
driven that pick. The table at the end of the report lists which species that
turns out to be, point by point.

```{r}
#| label: fig-weight-type1
#| fig-width: 12
#| fig-height: 5
plot_species_row(lapply(w_t1, as_rast), w_rng(w_t1),
                 "CV x ExDet Type 1 (phase 1 selection surface)", pal = magma_pal)
```

## CV × ExDet Type 2 — the phase 2 weighting

Only the combinationally novel pixels are candidates, so almost all of Kenya is
blank at iteration 0 — a scatter of pixels around Lake Turkana and the northern
rangelands. This is *not* why phase 2 turns out to be short: by the time phase 2
actually begins, phase 1 has cleared Type 1 and the Type-2 set is at its largest.
What ends phase 2 after two picks is the renormalisation the note below the
trajectories describes.

```{r}
#| label: fig-weight-type2
#| fig-width: 12
#| fig-height: 5
plot_species_row(lapply(w_t2, as_rast), w_rng(w_t2),
                 "CV x ExDet Type 2 (phase 2 selection surface)", pal = magma_pal)
```

## CV × Shape — the phase 3 weighting

Defined everywhere, so it never runs out. Phase 3 does *not* read this surface
over all of Kenya: it reads it only at the k-medoid cluster centres below.

```{r}
#| label: fig-weight-shape
#| fig-width: 12
#| fig-height: 5
plot_species_row(lapply(w_shape, as_rast), w_rng(w_shape),
                 "CV x Shape (phase 3 selection surface)", pal = magma_pal)
```

# The three-phase selection method

Each iteration chooses a single shared point and adds it to every species'
reference set, so all three evolve around common locations. Three features stay
constant across all phases:

1. **All species, every iteration.** We recompute MESS, ExDet and Shape for all
   three species on the current shared reference set, and combine the candidate
   surfaces across species with `pmax` — the most novel-and-uncertain species
   drives each cell.
2. **Weighted by uncertainty.** Every phase maximises a non-negative
   `CV × novelty` product, so the criterion can never flip sign — the failure
   mode of a `min(MESS × CV)` rule, which inverts to *avoid* uncertain pixels
   once nothing univariately novel remains.
3. **Distinct cells.** We mask out each pixel as we choose it, forcing a new
   location every iteration.

The phases differ in *which* novelty measure they maximise and *which pixels are
eligible*:

1. **Phase 1 — maximise `CV × Type-1`** (`−exdet` where `exdet < 0`) over all
   pixels. Runs until no `exdet < 0` pixel remains.
2. **Phase 2 — maximise `CV × Type-2`** (`exdet − 1` where `exdet > 1`) over all
   pixels. Runs until no `exdet > 1` pixel remains.
3. **Phase 3 — maximise `CV × Shape` over cluster medoids only.** We partition
   the landscape into `n_clust` environmental clusters with k-medoids on the
   standardised covariates. Each cluster contributes exactly one candidate — its
   medoid, the most typical land cell in that cluster. The phase repeatedly
   takes the unused medoid with the highest `CV × Shape` and adds it.

Phase 3 is the deliberate change of intent. Phases 1 and 2 chase extremes,
because extremes are what create range and combination extrapolation. Once those
run out, maximising `CV × Shape` pixel-by-pixel would send researchers to
whatever single pixel is furthest from the nearest existing record — which will
tend to be an environmental outlier representing little else in the country.
Restricting phase 3 to medoids replaces "the most extreme pixel" with "the most
representative pixel of an environment we haven't sampled": it bounds how
novel a chosen site can be, rather than maximising it, while still ranking the
candidates by uncertainty. (This report does not run the unrestricted-Shape
alternative, so the comparison is an argument for the design, not a measured
result.)

## Partitioning the landscape into k-medoid clusters

```{r}
#| label: kmedoid-partition

# Standardise over the whole landscape so Euclidean distances weight every
# covariate equally, then partition the land cells.
Z_land <- scale(values(cov_rast_ken_sub)[land_cells, , drop = FALSE])

km <- kmedoids(Z_land, k = n_clust, seed = 1)
medoid_cells <- land_cells[km$medoid]           # raster cell of each medoid

clust_rast <- rast(cov_rast_ken_sub[[1]])
values(clust_rast) <- NA_real_
values(clust_rast)[land_cells] <- km$cluster

# Clusters that already hold at least one existing record, pooled over species:
# a visited site is sampling effort whichever species were caught there, and the
# design is shared. Every other cluster is an environment type never sampled.
record_cells   <- unique(cellFromXY(cov_rast_ken_sub, crds(ken_points)))
record_cells   <- record_cells[!is.na(record_cells)]
sampled_clust  <- unique(km$cluster[match(record_cells, land_cells)])
sampled_clust  <- sampled_clust[!is.na(sampled_clust)]

# Binary derivative of clust_rast: 1 = cluster already sampled, 0 = gap, NA off
# land (`%in%` would turn the off-land NAs into 0, so carry them through).
clust_vals    <- values(clust_rast)[, 1]
sampled_rast  <- rast(clust_rast)
values(sampled_rast) <- ifelse(is.na(clust_vals), NA_real_,
                               as.numeric(clust_vals %in% sampled_clust))

cat(sprintf(
  "%d land cells partitioned into %d clusters (converged in %d iterations).\n",
  length(land_cells), n_clust, km$iters))
clust_sizes <- as.integer(table(km$cluster))
cat(sprintf("Cluster sizes: median %.0f cells, range %d-%d.\n",
            median(clust_sizes), min(clust_sizes), max(clust_sizes)))
cat(sprintf("%d of %d clusters already hold an existing record; %d are gaps.\n",
            length(sampled_clust), n_clust, n_clust - length(sampled_clust)))
```

Each cluster is one "type" of environment, and its medoid is the real land cell
that best represents it — the only site phase 3 can ever pick from that cluster.
The right-hand panel collapses the partition to what matters for the design:
white clusters already hold an existing record, black clusters are environment
types nobody has ever sampled.

The black area is the gap the design is working into, but note what phase 3 does
*not* do: it does not confine itself to that gap. Every unused medoid is a
candidate, so it can still pick a white cluster if the `CV × Shape` there is
high, and nothing stops two points landing in the same cluster. The end of the
report says how much of that happens in this run.

```{r}
#| label: fig-kmedoids
#| fig-width: 12
#| fig-height: 6.5
med_xy <- xyFromCell(cov_rast_ken_sub, medoid_cells)
# Bottom margin reserved on both panels for the legend: Kenya fills its panel,
# so a legend inside the map area clips against the panel edge.
par(mfrow = c(1, 2), mar = c(3, 1, 3, 1))
plot(clust_rast, col = sample(hcl.colors(n_clust, "Spectral")), legend = FALSE,
     axes = FALSE, main = sprintf("%d environmental clusters", n_clust))
lines(ken)

# Fixed breaks (not a continuous ramp) so 0 is always black and 1 always white.
plot(sampled_rast, breaks = c(-0.5, 0.5, 1.5), col = c("black", "white"),
     legend = FALSE, axes = FALSE,
     main = sprintf("%d of %d clusters already sampled",
                    length(sampled_clust), n_clust))
lines(ken, col = "grey60")
# Medoids in orange: legible on both the black and the white clusters.
points(med_xy, pch = 21, cex = 0.8, bg = phase_cols[["medoid"]], col = "grey30")
legend("bottom", inset = c(0, -0.06), xpd = NA, horiz = TRUE, bty = "n",
       cex = 0.9, legend = c("holds a record", "never sampled", "medoid"),
       pch = c(22, 22, 21), pt.cex = c(1.7, 1.7, 1),
       pt.bg = c("white", "black", phase_cols[["medoid"]]),
       col = c("grey40", "grey40", "grey30"))
```

# Running the selection

The loop below does everything in one pass. At each iteration it holds the
reference set (records + points chosen so far) and, for every species, computes
MESS, ExDet and Shape; records per-species per-iteration totals;
keeps every iteration's surface (the rasters are small); then selects and adds
the next shared point using the phase rule above.

```{r}
#| label: run-selection
#| code-fold: show
#| cache: false

grid_iters <- round(seq(0, n_points, length.out = 9))  # 9 iters shown in grids

ref_vals     <- ref_vals_0        # grows as points are added (shared)
chosen_cells <- integer(0)        # grid cells already selected (kept distinct)

selected_points <- data.frame(
  longitude = numeric(0), latitude = numeric(0), source_species = character(0),
  iteration = integer(0), phase = character(0), sel_value = numeric(0),
  cluster = integer(0)
)

# Per-iteration, per-species scalar trajectories.
traj_list <- vector("list", n_points + 1)

# Keep every iteration's surface, per species (small rasters). Indexed by
# iteration: [[sp]][[i + 1]] is the surface after i points have been added.
mess_surfaces  <- setNames(lapply(ken_mess_spp, \(s) vector("list", n_points + 1)), ken_mess_spp)
exdet_surfaces <- setNames(lapply(ken_mess_spp, \(s) vector("list", n_points + 1)), ken_mess_spp)
shape_surfaces <- setNames(lapply(ken_mess_spp, \(s) vector("list", n_points + 1)), ken_mess_spp)

for (i in 0:n_points) {

  # All three metrics for every species on the current shared reference set.
  mess_i  <- setNames(lapply(ref_vals, \(rv) mess(cov_rast_ken_sub, rv)$mess),   ken_mess_spp)
  exdet_i <- setNames(lapply(ref_vals, \(rv) exdet(cov_rast_ken_sub, rv)$exdet), ken_mess_spp)
  shape_i <- setNames(lapply(ref_vals, \(rv) shape_eval(rv)),                    ken_mess_spp)

  mess_vals  <- lapply(mess_i,  \(r) values(r)[, 1])
  exdet_vals <- lapply(exdet_i, \(r) values(r)[, 1])
  shape_vals <- lapply(shape_i, \(r) values(r)[, 1])

  traj_list[[i + 1]] <- data.frame(
    iteration   = i,
    species     = ken_mess_spp,
    mess_mean   = sapply(mess_vals,  \(v) mean(v, na.rm = TRUE)),
    exdet_type1 = sapply(exdet_vals, \(v) sum(-v[v < 0],    na.rm = TRUE)),
    exdet_type2 = sapply(exdet_vals, \(v) sum(v[v > 1] - 1, na.rm = TRUE)),
    shape_mean  = sapply(shape_vals, \(v) mean(v, na.rm = TRUE)),
    row.names   = NULL
  )

  for (sp in ken_mess_spp) {
    mess_surfaces[[sp]][[i + 1]]  <- mess_i[[sp]]
    exdet_surfaces[[sp]][[i + 1]] <- exdet_i[[sp]]
    shape_surfaces[[sp]][[i + 1]] <- shape_i[[sp]]
  }

  if (i == n_points) break

  # Phase rule: Type-1 anywhere, then Type-2 anywhere, then CV x Shape at the
  # unused cluster medoids. Each candidate surface is CV x novelty, reduced
  # across species by pmax, with already-chosen cells masked out.
  s1 <- lapply(ken_mess_spp, \(sp) cv_cols[[sp]] * type1_mag(exdet_vals[[sp]]))
  cell1 <- pmax_na(s1); cell1[chosen_cells] <- NA
  if (any(!is.na(cell1))) {
    phase <- "type1"; s_sel <- s1; cell <- cell1
  } else {
    s2 <- lapply(ken_mess_spp, \(sp) cv_cols[[sp]] * type2_mag(exdet_vals[[sp]]))
    cell2 <- pmax_na(s2); cell2[chosen_cells] <- NA
    if (any(!is.na(cell2))) {
      phase <- "type2"; s_sel <- s2; cell <- cell2
    } else {
      s3 <- lapply(ken_mess_spp, \(sp) cv_cols[[sp]] * shape_vals[[sp]])
      cell3 <- pmax_na(s3)
      # Phase 3 candidates: cluster medoids only, minus anything already taken.
      cand  <- setdiff(medoid_cells, chosen_cells)
      keep  <- rep(NA_real_, length(cell3)); keep[cand] <- cell3[cand]
      cell3 <- keep
      if (all(is.na(cell3))) { message("No medoid candidates left at iter ", i); break }
      phase <- "medoid"; s_sel <- s3; cell <- cell3
    }
  }

  sel_cell  <- which.max(cell)                                       # global pick
  sel_value <- cell[sel_cell]
  sp_name   <- ken_mess_spp[which.max(sapply(s_sel, \(v) v[sel_cell]))]
  sel_clust <- km$cluster[match(sel_cell, land_cells)]

  xy <- xyFromCell(cov_rast_ken_sub, sel_cell)
  selected_points <- rbind(selected_points, data.frame(
    longitude = xy[1, "x"], latitude = xy[1, "y"], source_species = sp_name,
    iteration = i, phase = phase, sel_value = sel_value, cluster = sel_clust
  ))
  chosen_cells <- c(chosen_cells, sel_cell)

  # Add the chosen point to every species' reference set.
  new_vals <- terra::extract(
    cov_rast_ken_sub,
    vect(data.frame(x = xy[1, "x"], y = xy[1, "y"]), geom = c("x", "y"),
         crs = crs(cov_rast_ken_sub)),
    ID = FALSE
  )
  for (sp in ken_mess_spp) ref_vals[[sp]] <- rbind(ref_vals[[sp]], new_vals)
}

traj <- do.call(rbind, traj_list)
phase_levels <- c("type1", "type2", "medoid")
selected_points$phase <- factor(selected_points$phase, levels = phase_levels)

n_by_phase <- table(selected_points$phase)
cat(sprintf("Selected %d distinct points.\n", nrow(selected_points)))
cat("Points per phase:\n"); print(n_by_phase)
```

# Results

## How many points per phase

Phase 1 (univariate novelty) absorbs the extreme, out-of-range environments
first; it exhausts once every pixel is univariately in range. Phase 2
(combinational novelty) ends when no pixel still qualifies as Type 2 — in this
run after two picks, because those picks inflate the Mahalanobis normaliser Type
2 divides by (see the note below the trajectories). Phase 3 then takes the
remainder of the budget, each iteration selecting the highest-scoring remaining
medoid; because we recompute Shape after every addition, that ranking changes
each time rather than standing fixed in advance.

```{r}
#| label: points-by-phase
tibble::enframe(n_by_phase, name = "phase", value = "points") |>
  mutate(phase = phase_labels[as.character(phase)],
         points = as.integer(points)) |>
  knitr::kable()
```

## The selected points

The deliverable: shared survey locations, with colour marking the phase that
chose each one. Phase 1 goes wherever the most extreme covariate values happen to
be, and those run the length of the country — the Turkana basin, the coastal
strip and inland sites alike — rather than clustering at one margin. Phase 3
points are cluster medoids, so they sit in the *body* of the unsampled
environments rather than at their edges. The full set — selection order, source
species, phase, cluster and the `CV × novelty` value at selection — goes to
`k_m_r_selected_points.csv` in `kenya_mess/`, and the table at the end of the
report prints it in full.

```{r}
#| label: fig-selected-points
#| fig-width: 7
#| fig-height: 8
plot(ken, main = sprintf("%d selected survey points", nrow(selected_points)))
points(vect(selected_points, geom = c("longitude", "latitude"), crs = crs(cov_rast_ken_sub)),
       pch = 19, cex = 1, col = phase_cols[as.character(selected_points$phase)])
legend("bottomleft", legend = phase_labels[phase_levels], pch = 19,
       col = phase_cols[phase_levels], bty = "n", title = "selected in", inset = 0.02)
```

```{r}
#| label: export-points

# Shape at each selected point at the moment it was selected: how novel the site
# a researcher is being sent to actually is. Quoted in the trajectory section and
# tabulated at the end of the report.
shape_at_pt <- sapply(seq_len(nrow(selected_points)), \(j) {
  cell <- cellFromXY(cov_rast_ken_sub,
                     as.matrix(selected_points[j, c("longitude", "latitude")]))
  max(sapply(ken_mess_spp, \(sp)
             values(shape_surfaces[[sp]][[selected_points$iteration[j] + 1]])[cell, 1]))
})
selected_points$shape_at_selection <- shape_at_pt

# Per-phase spread, quoted in the text later.
sh_get <- function(ph, f) {
  v <- selected_points$shape_at_selection[selected_points$phase == ph]
  if (!length(v)) NA_integer_ else round(f(v))
}

selected_points <- selected_points |>
  mutate(priority = row_number(), .before = 1)
write_csv(selected_points, "k_m_r_selected_points.csv")
```

# Evolution of the novelty metrics

## Trajectories

Three of the four panels show the design working: mean MESS rises (fewer pixels
are univariately out of range), ExDet Type 1 collapses to zero, and mean Shape
declines. ExDet Type 2 moves the other way first — it is close to zero at
iteration 0, grows by more than an order of magnitude through phase 1, and only
then collapses. That is the masking rule at work, not the design creating
novelty; see the note below. Note also that the sharp fall in Shape coincides
with the two Type-2 picks rather than with any part of phase 1. One line per
species; dashed lines mark the phase boundaries (end of Type 1, end of Type 2).

```{r}
#| label: fig-trajectories
#| fig-width: 10
#| fig-height: 7
metric_levels <- c("MESS (mean over cells)", "ExDet Type 1 (total, univariate)",
                   "ExDet Type 2 (total, combinational)",
                   "Shape (mean over cells)")
traj_long <- traj |>
  transmute(iteration, species,
            `MESS (mean over cells)`             = mess_mean,
            `ExDet Type 1 (total, univariate)`   = exdet_type1,
            `ExDet Type 2 (total, combinational)`= exdet_type2,
            `Shape (mean over cells)`            = shape_mean) |>
  pivot_longer(-c(iteration, species), names_to = "metric", values_to = "value") |>
  mutate(metric  = factor(metric, levels = metric_levels),
         species = factor(species, levels = ken_mess_spp))

n_type1 <- sum(selected_points$phase == "type1")
n_type2 <- sum(selected_points$phase == "type2")
phase_bounds <- c(n_type1, n_type1 + n_type2)
phase_bounds <- phase_bounds[phase_bounds > 0 & phase_bounds < n_points]

ggplot(traj_long, aes(iteration, value, colour = species)) +
  geom_vline(xintercept = phase_bounds, linetype = "dashed",
             colour = "grey55", linewidth = 0.5) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = setNames(
    c("#D6604D", "#4393C3", "#1B7837"), ken_mess_spp)) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(x = "Iteration (points added)", y = "Value",
       title = "Novelty metrics as the three-phase points are added") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

Reading that decline at the individual sites rather than over the map shows what
phase 3 is for. Phase 1 takes whatever is most extreme at the time, so the Shape
value at its picks, at the moment we picked them, spans ≈ `r sh_get("type1", min)`
to ≈ `r format(sh_get("type1", max), big.mark = ",")` — a couple of sites tens of
times more extrapolated than the typical spread of the records, others barely
novel at all. Every phase-3 medoid pick falls between ≈ `r sh_get("medoid", min)`
and ≈ `r sh_get("medoid", max)`. The medians are much closer together than that
contrast suggests (`r sh_get("type1", median)` for phase 1 against
`r sh_get("medoid", median)` for phase 3), so the trade is not that later sites
are uniformly less novel — it is that phase 3 *cannot* send anyone to an extreme
outlier.

::: {.callout-note collapse="true"}
## Why ExDet Type 2 climbs steeply before it collapses (an artefact, not created novelty)

A pixel reports its ExDet score as `ifelse(nt1 == 0, nt2, nt1)`, so while a pixel
is univariately novel its Type 1 score masks its combinational novelty. Almost
all of Kenya starts out Type 1, which is why the Type-2 area looks negligible at
iteration 0; as phase 1 strips Type 1 away, those pixels reveal Type 2 values
they had all along, and the sum climbs by more than an order of magnitude. The
design creates none of that novelty — ExDet simply was not reporting it.

On top of that, `nt2 = mahalanobis / mah_max` divides by the *most-extreme
reference point*, so adding an extreme point inflates `mah_max` and rescales the
whole surface. That is what ends phase 2: two Type-2 picks move the normaliser
far enough that no pixel exceeds the threshold, and Type 2 drops to exactly zero
for the rest of the run. So the shortness of phase 2 is a property of the metric,
not evidence that the design has resolved the combinational novelty.
:::

## Surface evolution, species by species

Iteration 0, iteration `r n_points`, and 7 evenly-spaced steps between; each panel
overlays the points chosen so far (dark = first selected, yellow = last). One
3 × 3 grid per species, for each of the three metrics.

### MESS

Univariate novelty (red) gives way to similarity (blue) as phase 1 absorbs the
extreme environments.

```{r}
#| label: fig-mess-evolution
#| fig-width: 10
#| fig-height: 11
mess_range_all <- clamp_range(unlist(
  lapply(ken_mess_spp, \(sp) mess_surfaces[[sp]][grid_iters + 1]), recursive = FALSE))
for (sp in ken_mess_spp) {
  plot_mess_evolution(
    surfaces   = mess_surfaces[[sp]][grid_iters + 1],
    iters      = grid_iters,
    pts        = selected_points,
    mess_range = mess_range_all,
    title      = bquote(italic(.(sp)) * " — MESS over the selected points")
  )
}
```

### ExDet

Type 1 (red) covers nearly the whole country at iteration 0. It shrinks steadily
through phase 1 but is still extensive at the first grid step and only clears
around iteration 12. Type 2 (purple) runs in the opposite direction: there is
essentially none at iteration 0, a large north-western block appears as the
Type-1 mask lifts from those pixels, and it then disappears completely between
the iteration-19 and iteration-25 panels, when the two Type-2 picks land inside
it. From there on ExDet classes every pixel "similar" for all three species.

```{r}
#| label: fig-exdet-evolution
#| fig-width: 10
#| fig-height: 11
for (sp in ken_mess_spp) {
  plot_exdet_evolution(
    surfaces = exdet_surfaces[[sp]][grid_iters + 1],
    iters    = grid_iters,
    pts      = selected_points,
    title    = bquote(italic(.(sp)) * " — ExDet over the selected points")
  )
}
```

### Shape

The bright north-western block at iteration 0 is the most extrapolated region for
all three species, and it survives the whole of phase 1 — it is still there in
the iteration-19 panel, after we have placed twenty points. It vanishes between
iterations 19 and 25, when the two Type-2 points land inside it: because Shape
measures distance to the *nearest* record, two records in the middle of a novel
region collapse the whole region at once. After that the surface darkens slowly
as the medoid points fill the remaining gaps.

All nine panels share one colour scale, set by that iteration-0 extreme, so it
compresses late-iteration structure into the dark end — read the *Shape* panel of
the trajectory figure above for what is still changing after iteration 25.

```{r}
#| label: fig-shape-evolution
#| fig-width: 10
#| fig-height: 11
shape_range <- c(0, as.numeric(quantile(unlist(lapply(
  ken_mess_spp, \(sp) lapply(shape_surfaces[[sp]][grid_iters + 1], \(r) values(r)[, 1])
)), 0.99, na.rm = TRUE)))
for (sp in ken_mess_spp) {
  plot_shape_evolution(
    surfaces    = shape_surfaces[[sp]][grid_iters + 1],
    iters       = grid_iters,
    pts         = selected_points,
    shape_range = shape_range,
    title       = bquote(italic(.(sp)) * " — Shape over the selected points")
  )
}
```


# What the design covers

## Environment types with any sampling

The same black-and-white view of the `r n_clust` environmental clusters as
before, but now counting both kinds of point: white clusters hold at least
one existing record *or* at least one newly selected site, black clusters hold
neither and remain entirely unsampled. The map overlays the selected points, so
the black-to-white conversions the design achieves show directly.

```{r}
#| label: fig-cluster-coverage
#| fig-width: 7
#| fig-height: 8

sel_clusters  <- stats::na.omit(selected_points$cluster)
covered_after <- sort(unique(c(sampled_clust, sel_clusters)))
new_clusters  <- setdiff(covered_after, sampled_clust)

# Same binary construction as the earlier partition figure: fixed breaks so 0 is
# always black and 1 always white, NA carried through off-land.
covered_rast <- rast(clust_rast)
values(covered_rast) <- ifelse(is.na(clust_vals), NA_real_,
                               as.numeric(clust_vals %in% covered_after))

par(mar = c(3, 1, 3, 1))
plot(covered_rast, breaks = c(-0.5, 0.5, 1.5), col = c("black", "white"),
     legend = FALSE, axes = FALSE,
     main = sprintf("%d of %d clusters hold a record or a selected point",
                    length(covered_after), n_clust))
lines(ken, col = "grey60")
points(vect(selected_points, geom = c("longitude", "latitude"),
            crs = crs(cov_rast_ken_sub)),
       pch = 21, cex = 0.9, bg = phase_cols[["medoid"]], col = "grey30")
legend("bottom", inset = c(0, -0.04), xpd = NA, horiz = TRUE, bty = "n",
       cex = 0.9, legend = c("has a point", "still unsampled", "selected site"),
       pch = c(22, 22, 21), pt.cex = c(1.7, 1.7, 1),
       pt.bg = c("white", "black", phase_cols[["medoid"]]),
       col = c("grey40", "grey40", "grey30"))

n_repeat  <- sum(duplicated(sel_clusters))
n_already <- sum(selected_points$cluster[selected_points$phase == "medoid"]
                 %in% sampled_clust)
```

Before the design, `r length(sampled_clust)` of the `r n_clust` clusters held a
record. The `r nrow(selected_points)` selected points land in
`r length(unique(sel_clusters))` distinct clusters, of which
`r length(new_clusters)` were previously unsampled, bringing coverage to
`r length(covered_after)` of `r n_clust`.

Two costs of not conditioning the selection on cluster coverage show up here.
`r n_repeat` points fall in a cluster another selected point already occupies,
and `r n_already` of the phase-3 medoid picks sit in a cluster that already held
an existing record. Phase 3 allows both, because it ranks *every* unused medoid
by `CV × Shape` and never asks whether anyone has visited that environment.
Excluding already-covered clusters from the phase-3 candidate pool would convert
those picks into new coverage, at the cost of ignoring uncertainty in
environments that are nominally sampled but thin on records.

## Which species drove each pick

Each iteration takes the cross-species maximum, so exactly one species sets the
score at the winning pixel: that is the species whose extrapolation and
uncertainty justified the site. The full selection order follows.

```{r}
#| label: tbl-drivers
drv <- table(factor(selected_points$source_species, levels = ken_mess_spp))
never_drv <- names(drv)[drv == 0]

cat("Picks driven by each species:\n"); print(drv)

selected_points |>
  transmute(priority, iteration, phase,
            `driven by` = source_species,
            longitude, latitude,
            `CV x novelty` = sel_value,
            `Shape at site` = shape_at_selection) |>
  knitr::kable(digits = 3, row.names = FALSE)
```

```{r}
#| label: drivers-note
#| output: asis
if (length(never_drv)) {
  cat(sprintf(
    "Note that *%s* never drives a pick: at every selected pixel another species is at least as extrapolated and uncertain. The shared design still adds each point to its reference set, so it benefits from the coverage without ever setting the criterion.\n",
    paste(never_drv, collapse = "*, *")))
}
```
