---
title: "Kenya MESS"
format: html
---

## Setup

### Data
```{r}
#| label: setup
#| message: false

library(targets)
library(terra)
library(dplyr)
library(tidyr)
library(ggplot2)
library(bssdm)

# spatial data 
tar_load(
  c(
    covariate_rast_5,
    model_data_spatial,
    ken,
    target_covariate_names
  ),
  store = "~/Documents/tki_work/vector_atlas/va_multispecies_sdm/_targets/"
)

# 
cov_rast_ken <- crop(covariate_rast_5, ken) |>
  mask(ken)

cov_rast_ken_sub <- cov_rast_ken[[target_covariate_names]]

# species for this analysis
ken_mess_spp <- c("arabiensis", "gambiae", "funestus")

# coefficient of variation
cv <- rast("~/Documents/tki_work/vector_atlas/va_multispecies_sdm/spartan_model_comparison/m6/m6_mask_p_cv.tif") |> 
  crop(ken) |> 
  mask(ken)

cv <- cv[[ken_mess_spp]]

# Resample CV to match the covariate raster resolution for later weighting
cv <- resample(cv, cov_rast_ken_sub, method = "bilinear")

# Muted divergent shades used throughout (negative/novel vs positive/similar)
col_neg <- "#D6604D"
col_pos <- "#4393C3"
```

### Look at some of the data 

#### Environmental Covariates

```{r}
#| label: cov-rast-ken
par(mfrow = c (2, 5))
for(i in 1:10) {
  plot(
    x = cov_rast_ken_sub[[i]],
    legend = FALSE,
    axes = FALSE,
    main = names(cov_rast_ken_sub)[i]
  )
  lines(ken)
}
```

#### Existing vector atlas sample points

```{r}
#| label: point-data

model_data_vect <- vect(
  model_data_spatial |>
    select(
      species,
      longitude,
      latitude,
      count,
      presence,
      data_type
    ) |> 
    filter(species %in% ken_mess_spp),
  geom = c("longitude", "latitude"),
  crs = crs(ken)
)


ken_points <- model_data_vect[ken, ]

par(mfrow = c(1, 3), mar = c(2, 2, 2, 1))

for (sp in ken_mess_spp) {
  sp_pts <- ken_points[ken_points$species == sp, ]
  plot(ken)
  title(main = bquote(italic(.(sp))))
  points(sp_pts, pch = 16, col = col_pos, cex = 0.8)
}
```

#### Distribution uncertainty - coefficient of variation

```{r}
#| label: cv-plot
#| fig-width: 12
#| fig-height: 5

# Calculate shared range across all species (symmetric for comparison)
cv_range <- range(unlist(lapply(ken_mess_spp, function(sp) {
  quantile(values(cv[[sp]]), c(0.01, 0.99), na.rm = TRUE)
})), na.rm = TRUE)

# Layout: three map panels above one shared colorbar
layout(rbind(1:3, c(4, 4, 4)), heights = c(1, 0.28))
par(mar = c(2, 2, 3, 1))

for (sp in ken_mess_spp) {
  plot(
    cv[[sp]],
    col = map.pal("magma", 100) |> 
      rev(x = _),
    range = cv_range,
    legend = FALSE,
    axes = FALSE
  )
  title(main = bquote(italic(.(sp))))
  lines(ken)
}

# Shared colorbar
par(mar = c(3, 6, 0.5, 6))
magma_pal <- map.pal("magma", 100)
cv_vals <- seq(cv_range[1], cv_range[2], length.out = length(magma_pal))
image(x = cv_vals, y = 1, z = matrix(cv_vals, ncol = 1),
      col = magma_pal, axes = FALSE, xlab = "Coefficient of Variation (CV)", ylab = "")
axis(1)
```


## Multivariate Environmental Similarity Surface

The Multivariate Environmental Similarity Surface (MESS) is a method for quantifying how similar environmental conditions at each location are to the reference environmental space defined by observation points. For each pixel in a map, MESS compares its covariate values to the range and distribution of values observed in the reference set (e.g., species occurrences or survey points).

- **Positive MESS values** (blue in maps): The environment is similar to the reference set. Higher values indicate greater similarity—the location falls well within the observed covariate ranges.
- **Negative MESS values** (red in maps): The environment is novel or extrapolated. The location falls *outside* the range of at least one covariate observed in the reference set, indicating potential extrapolation risk for models trained on the reference data.
- **Zero or near-zero**: (white in maps) The location is at or near the edge of the observed environmental space.


###  MESS by species

Here MESS surfaces are calculated for each species separately using their respective observation points as the reference set:

```{r}
#| label: mess-iter0

# Reference env values per species (observation points only)
ref_vals_0 <- lapply(ken_mess_spp, function(sp) {
  sp_points <- ken_points[ken_points$species == sp, ]
  terra::extract(cov_rast_ken_sub, sp_points, ID = FALSE) |>
    na.omit()
})
names(ref_vals_0) <- ken_mess_spp

# Iteration 0 MESS surface per species
mess_iter0 <- lapply(ref_vals_0, function(rv) mess(cov_rast_ken_sub, rv))
names(mess_iter0) <- ken_mess_spp

# extract just the mess rasters
iter0 <- lapply(mess_iter0, \(m) m$mess)

# Shared symmetric range, clamped to the 1st/99th percentile so extreme outliers
# don't wash out the colour variation. Zero maps to white for comparability.
iter0_absmax <- max(abs(unlist(lapply(iter0, \(r) {
  quantile(values(r), c(0.01, 0.99), na.rm = TRUE)
}))))
iter0_range  <- c(-iter0_absmax, iter0_absmax)
divergent_pal <- colorRampPalette(c(col_neg, "white", col_pos))(255)

# Layout: three map panels above one shared horizontal colourbar
layout(rbind(1:3, c(4, 4, 4)), heights = c(1, 0.28))
par(mar = c(2, 2, 3, 1))
for (sp in ken_mess_spp) {
  plot(iter0[[sp]], col = divergent_pal, range = iter0_range, legend = FALSE, axes = FALSE)
  title(main = bquote(italic(.(sp))))
  lines(ken)
}

# Shared colourbar
par(mar = c(3, 6, 0.5, 6))
iter0_vals <- seq(iter0_range[1], iter0_range[2], length.out = length(divergent_pal))
image(x = iter0_vals, y = 1, z = matrix(iter0_vals, ncol = 1),
      col = divergent_pal, axes = FALSE, xlab = "MESS (iteration 0)", ylab = "")
axis(1)
```

#### Most Dissimilar Variable (MOD)

While MESS identifies *whether* a location is novel by finding the covariate that most deviates from the reference range, the Most Dissimilar variable (MOD) identifies *which* covariate is most responsible for that novelty. For each pixel, MOD shows the index of the covariate with the maximum standardized deviation from its reference range—in other words, the covariate most "out of place" at that location.

```{r}
#| label: plot-iter0-mod
#| fig-width: 12
#| fig-height: 5

# Extract MOD (most dissimilar variable) for each species
# MOD is the index of the variable causing the most dissimilarity
iter0_mod <- lapply(mess_iter0, \(m) m$mod)
names(iter0_mod) <- ken_mess_spp

# Define color palette matching each covariate
mod_pal <- c(
  "#2ECC21", "#FF4081", "#F8F32B", "#A8DADC", "#457B9D",
  "#1D3557", "#FF5733", "#F1C40F", "#6C757D", "#C4E538"
)

n_covariates <- nlyr(cov_rast_ken_sub)

# Layout: three map panels above one shared legend panel
layout(rbind(1:3, c(4, 4, 4)), heights = c(1, 0.5))
par(mar = c(2, 2, 3, 1))

for (sp in ken_mess_spp) {
  plot(iter0_mod[[sp]], col = mod_pal, legend = FALSE, axes = FALSE, type = "classes")
  title(main = bquote(italic(.(sp))))
  lines(ken)
}

# Create a legible legend showing covariate names and colors
par(mar = c(1, 1, 1, 1))
plot(NA, xlim = c(0, 10), ylim = c(0, 10), axes = FALSE, xlab = "", ylab = "")

# Arrange legend in two columns for readability
col_1_idx <- seq_len(5)
col_2_idx <- seq(6, 10)
y_pos_1 <- seq(9, 1, by = -1.6)
y_pos_2 <- seq(9, 1, by = -1.6)

# First column (covariates 1-5)
for (i in seq_along(col_1_idx)) {
  cov_idx <- col_1_idx[i]
  rect(0.2, y_pos_1[i] - 0.4, 0.8, y_pos_1[i] + 0.4, 
       col = mod_pal[cov_idx], border = "black", lwd = 2)
  text(1.5, y_pos_1[i], paste0(cov_idx, ". ", target_covariate_names[cov_idx]), 
       cex = 0.95, adj = c(0, 0.5), font = 1)
}

# Second column (covariates 6-10)
for (i in seq_along(col_2_idx)) {
  cov_idx <- col_2_idx[i]
  rect(5.2, y_pos_2[i] - 0.4, 5.8, y_pos_2[i] + 0.4, 
       col = mod_pal[cov_idx], border = "black", lwd = 2)
  text(6.5, y_pos_2[i], paste0(cov_idx, ". ", target_covariate_names[cov_idx]), 
       cex = 0.95, adj = c(0, 0.5), font = 1)
}
```

#### Pairwise MESS differences between species

The MESS analyses for each species all look simlar. Let's check on that.

Each panel maps the cell-by-cell difference in MESS between pairs of
species.

```{r}
#| label: plot-iter0-diff
#| fig-width: 12
#| fig-height: 5

iter0_diffs <- list(
  "arabiensis - gambiae"  = iter0$arabiensis - iter0$gambiae,
  "arabiensis - funestus" = iter0$arabiensis - iter0$funestus,
  "gambiae - funestus"    = iter0$gambiae    - iter0$funestus
)

diff_absmax <- max(abs(unlist(lapply(iter0_diffs, \(r) minmax(r)))))
diff_range  <- c(-diff_absmax, diff_absmax)

# Distinct divergent palette (orange-teal) for the difference maps, symmetric
# around zero so no difference maps to white.
diff_pal <- colorRampPalette(c("#E08214", "white", "#80CDC1"))(255)

# Layout: three map panels above one shared horizontal colourbar
layout(rbind(1:3, c(4, 4, 4)), heights = c(1, 0.28))
par(mar = c(2, 2, 3, 1))
for (nm in names(iter0_diffs)) {
  plot(iter0_diffs[[nm]], col = diff_pal, range = diff_range,
       legend = FALSE, axes = FALSE)
  # Parse the pair names to italicize species names
  pair_parts <- strsplit(nm, " - ")[[1]]
  title(main = bquote(italic(.(pair_parts[1])) * " - " * italic(.(pair_parts[2]))))
  lines(ken)
}

# Shared colourbar
par(mar = c(3, 6, 0.5, 6))
diff_vals <- seq(diff_range[1], diff_range[2], length.out = length(diff_pal))
image(x = diff_vals, y = 1, z = matrix(diff_vals, ncol = 1),
      col = diff_pal, axes = FALSE, xlab = "Difference in iteration 0 MESS", ylab = "")
axis(1)
```

##### Novel (negative) regions -- are they similar across species?

Mask each surface to keep only the negative (novel / extrapolation)
cells, then compare where the novel regions fall across species.

```{r}
#| label: iter0-negative-similarity

# Keep only negative cells (novel areas)
iter0_neg <- lapply(iter0, \(r) { r[r >= 0] <- NA; r })

# Which cells are novel (negative) per species?
is_neg <- lapply(iter0, \(r) values(r) < 0)

# Pairwise agreement in the *location* of novel cells (Jaccard-style overlap)
overlap <- function(a, b) {
  both <- a & b
  either <- a | b
  sum(both, na.rm = TRUE) / sum(either, na.rm = TRUE)
}

cat("Overlap of novel locations (intersection / union):\n",
  "arabiensis vs gambiae: ", round(overlap(is_neg$arabiensis, is_neg$gambiae),  3), "\n",
  "arabiensis vs funestus:", round(overlap(is_neg$arabiensis, is_neg$funestus), 3), "\n",
  "gambiae vs funestus:   ", round(overlap(is_neg$gambiae,    is_neg$funestus), 3), "\n"
)

# Correlation of the negative MESS values where both species are novel
neg_cor <- function(a, b) {
  va <- values(a)[, 1]; vb <- values(b)[, 1]
  ok <- !is.na(va) & !is.na(vb)
  if (sum(ok) < 2) return(NA_real_)
  cor(va[ok], vb[ok])
}

cat("\nCorrelation of negative MESS values where both are novel:\n",
  "arabiensis vs gambiae: ", round(neg_cor(iter0_neg$arabiensis, iter0_neg$gambiae),  3), "\n",
  "arabiensis vs funestus:", round(neg_cor(iter0_neg$arabiensis, iter0_neg$funestus), 3), "\n",
  "gambiae vs funestus:   ", round(neg_cor(iter0_neg$gambiae,    iter0_neg$funestus), 3), "\n"
)
```

## MESS surfaces with iteratively added new sites

For each of the three target species

- iteratively generate a MESS surface (starting from the iteration 0 surface above),
- identify the cell with the lowest similarity score,
- record its centroid, add it to the point set, and
- repeat until an additional 20 points have been recorded.

```{r}
#| label: mess-iteration

n_iter <- 20
mess_results <- list()

for (sp in ken_mess_spp) {

  # Start from the species observation points
  ref_vals <- ref_vals_0[[sp]]

  recorded_pts <- data.frame(longitude = numeric(0), latitude = numeric(0))
  # Surfaces indexed 1..(n_iter + 1) for iterations 0..n_iter; iteration 0 is
  # the MESS before any search points have been added
  mess_surfaces <- vector("list", n_iter + 1)

  for (i in 0:n_iter) {

    # Reuse the pre-computed iteration 0 surface; otherwise recompute
    m <- if (i == 0) mess_iter0[[sp]] else mess(cov_rast_ken_sub, ref_vals)
    mess_surfaces[[i + 1]] <- m

    # Final surface computed; no further points to add
    if (i == n_iter) break

    # Identify cell with lowest MESS value and get its centroid
    min_cell <- which.min(values(m$mess))
    centroid  <- xyFromCell(m$mess, min_cell)
    lon <- centroid[1, "x"]
    lat <- centroid[1, "y"]

    recorded_pts <- rbind(
      recorded_pts,
      data.frame(longitude = lon, latitude = lat)
    )

    # Extract env values at new point and add to reference set
    new_pt   <- vect(data.frame(x = lon, y = lat), geom = c("x", "y"), crs = crs(cov_rast_ken_sub))
    new_vals <- terra::extract(cov_rast_ken_sub, new_pt, ID = FALSE)
    ref_vals <- rbind(ref_vals, new_vals)
  }

  mess_results[[sp]] <- list(
    mess_surfaces = mess_surfaces,
    recorded_points = recorded_pts
  )

  message("Done: ", sp)
}
```

## Points selected

```{r}
#| label: results

for (sp in ken_mess_spp) {
  cat("\n###", sp, "\n\n")
  print(mess_results[[sp]]$recorded_points)
}
```

## MESS surface evolution by species

Plot the MESS initially and the first eight iterations of MESS as the first 8 points are selected.

```{r}
#| label: plot-mess
#| fig-width: 10
#| fig-height: 11

# Clamp to the 1st/99th percentile across all surfaces (symmetric around zero)
# so extreme negative outliers don't wash out the colour variation. Zero maps
# to white and all panels share the same scale for comparability.
clamp_absmax <- max(abs(unlist(lapply(ken_mess_spp, function(sp) {
  lapply(mess_results[[sp]]$mess_surfaces, function(m) {
    quantile(values(m$mess), c(0.01, 0.99), na.rm = TRUE)
  })
}))))

mess_range <- c(-clamp_absmax, clamp_absmax)

# Divergent palette: muted red = negative (novel), white = zero, blue = positive
divergent_pal <- colorRampPalette(c(col_neg, "white", col_pos))(255)

# Layout: 3x3 grid of panels plus a bottom strip for the shared legend
layout_mat <- rbind(
  matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE),
  rep(10, 3)
)

for (sp in ken_mess_spp) {
  pts      <- mess_results[[sp]]$recorded_points
  surfaces <- mess_results[[sp]]$mess_surfaces

  layout(layout_mat, heights = c(1, 1, 1, 0.4))
  par(mar = c(1, 1, 2, 0.5), oma = c(0, 0, 3, 0))

  # Iterations 0 through 8 (surfaces 1 through 9)
  for (j in seq_len(9)) {
    iter <- j - 1
    plot(
      surfaces[[j]]$mess,
      main   = ifelse(
        iter == 0,
        "initial", paste0("iter ", iter)
      ),
      col    = divergent_pal,
      range  = mess_range,
      legend = FALSE,
      axes   = FALSE
    )
    lines(ken)

    # Show all points selected up to and including this iteration
    if (iter >= 1) {
      pts_so_far <- pts[seq_len(iter), , drop = FALSE]
      pts_vect   <- vect(pts_so_far, geom = c("longitude", "latitude"), crs = crs(cov_rast_ken_sub))
      points(pts_vect, pch = 19, col = "goldenrod", cex = 2)
    }
  }

  # Shared legend (panel 10): horizontal colour bar spanning the grid width
  par(mar = c(2.5, 3, 0.5, 3))
  legend_vals <- seq(mess_range[1], mess_range[2], length.out = length(divergent_pal))
  image(
    x = legend_vals, y = 1, z = matrix(legend_vals, ncol = 1),
    col = divergent_pal, axes = FALSE, xlab = "MESS (clamped to 1st/99th pctile)", ylab = ""
  )
  axis(1)

  mtext(bquote(italic(.(sp))), outer = TRUE, cex = 1.4, font = 2)
}
```

## These looks similar. Are the points actually identical?

```{r}
#| label: selected-points-identical

# Check whether the selected points are identical across the three species
selected_pts <- lapply(mess_results, \(x) x$recorded_points)

cat(
  "arabiensis vs gambiae: ", identical(selected_pts$arabiensis, selected_pts$gambiae),  "\n",
  "arabiensis vs funestus:", identical(selected_pts$arabiensis, selected_pts$funestus), "\n",
  "gambiae vs funestus:   ", identical(selected_pts$gambiae,    selected_pts$funestus), "\n",
  "all three identical:   ",
  identical(selected_pts$arabiensis, selected_pts$gambiae) &&
    identical(selected_pts$arabiensis, selected_pts$funestus), "\n"
)
```


## Are species observation points identical?

```{r}
#| label: points-identity

coords_by_species <- lapply(ken_mess_spp, function(sp) {
  pts <- crds(ken_points[ken_points$species == sp, ])
  sort(paste(round(pts[, "x"], 6), round(pts[, "y"], 6)))
})
names(coords_by_species) <- ken_mess_spp

# Point counts per species
sapply(coords_by_species, length)

# Pairwise identity checks
cat(
  "arabiensis vs gambiae: ", identical(coords_by_species[["arabiensis"]], coords_by_species[["gambiae"]]), "\n",
  "arabiensis vs funestus:", identical(coords_by_species[["arabiensis"]], coords_by_species[["funestus"]]), "\n",
  "gambiae vs funestus:   ", identical(coords_by_species[["gambiae"]],    coords_by_species[["funestus"]]),  "\n"
)
```

## Evolution of MESS surface

What is the sum of novel (negative), and similar (positive) values across the 
surface as MESS iterates? 

For each surface (from iteration 0), sum the positive cell values (similar,
blue) and the absolute value of the negative cells (novel, red), then track how
these totals change across iterations. The grey line (secondary axis) shows the
MESS value of the point selected at each iteration -- i.e. how novel the most
novel cell is -- so we can see what is being picked as the search continues.

```{r}
#| label: mess-cell-sums
#| fig-width: 12
#| fig-height: 4

cell_sums <- lapply(ken_mess_spp, function(sp) {
  surfaces <- mess_results[[sp]]$mess_surfaces
  data.frame(
    species   = sp,
    iteration = seq_along(surfaces) - 1,
    positive  = sapply(surfaces, function(m) {
      v <- values(m$mess); sum(v[v > 0], na.rm = TRUE)
    }),
    negative  = sapply(surfaces, function(m) {
      v <- values(m$mess); sum(abs(v[v < 0]), na.rm = TRUE)
    })
  )
}) |>
  bind_rows() |>
  pivot_longer(c(positive, negative), names_to = "sign", values_to = "abs_sum")

# MESS value of the point selected at each iteration (the most novel cell).
# Points are selected at iterations 0..(n_iter - 1); the final surface has none.
selected_mess <- lapply(ken_mess_spp, function(sp) {
  surfaces <- mess_results[[sp]]$mess_surfaces
  data.frame(
    species    = sp,
    iteration  = seq_len(n_iter) - 1,
    mess_value = sapply(surfaces[seq_len(n_iter)], function(m) {
      min(values(m$mess), na.rm = TRUE)
    })
  )
}) |>
  bind_rows()

# Combine into one long frame with panels for: absolute sums, selected-point MESS
# value, and iteration-to-iteration change in novel cells (each with a free y-axis).
plot_data <- bind_rows(
  cell_sums |>
    transmute(species, iteration, series = sign, value = abs_sum,
              panel = "Absolute sum of MESS cells"),
  selected_mess |>
    transmute(species, iteration, series = "selected", value = mess_value,
              panel = "Selected point MESS value"),
  cell_sums |>
    pivot_wider(names_from = sign, values_from = abs_sum) |>
    group_by(species) |>
    mutate(delta_negative = negative - lag(negative)) |>
    ungroup() |>
    transmute(species, iteration, series = "delta", value = delta_negative,
              panel = "Δ|negative| (novelty reduction)")
) |>
  mutate(panel = factor(
    panel,
    levels = c("Absolute sum of MESS cells", "Selected point MESS value", "Δ|negative| (novelty reduction)")
  ))

ggplot(plot_data, aes(iteration, value, colour = series)) +
  geom_line(data = ~ filter(.x, series %in% c("selected", "delta"))) +
  geom_point() +
  facet_grid(panel ~ species, scales = "free_y", switch = "y") +
  scale_colour_manual(
    values = c(negative = col_neg, positive = col_pos, selected = "grey40", delta = "steelblue"),
    labels = c(
      negative = "|negative| (novel)",
      positive = "positive (similar)",
      selected = "selected point MESS",
      delta = "Δ|negative|"
    ),
    breaks = c("negative", "positive", "selected", "delta")
  ) +
  labs(x = "Iteration", y = NULL, colour = NULL) +
  theme(strip.placement = "outside")
```

## Same data in a table

```{r}
#| label: mess-cell-sums-table

cell_sums |>
  pivot_wider(names_from = sign, values_from = abs_sum) |>
  left_join(selected_mess, by = c("species", "iteration")) |>
  arrange(species, iteration) |>
  group_by(species) |>
  mutate(delta_negative = negative - lag(negative)) |>
  ungroup() |>
  knitr::kable(
    digits    = c(0, 0, 0, 0, 2, 0),
    col.names = c("Species", "Iteration", "positive (similar)", "|negative| (novel)", "Selected point MESS", "Δ|negative|")
  )
```

So returns diminish, but, not linearly.


# Combined MESS-only point selection (all species)

Here the iteration selects a **single location per iteration** by computing MESS for each species, then picking the cell with the *lowest* MESS value **across all species**. Unlike the CV-weighted combined approach, this relies purely on MESS (novelty/extrapolation) without covariate diversity weighting. The selected location is added to all three species' reference sets.

```{r}
#| label: mess-iteration-combined

ref_vals_combined_mess <- ref_vals_0
recorded_pts_combined_mess <- data.frame(longitude = numeric(0), latitude = numeric(0))
mess_surfaces_combined_mess <- setNames(
  lapply(ken_mess_spp, \(sp) vector("list", n_iter + 1)),
  ken_mess_spp
)

for (i in 0:n_iter) {

  # MESS for every species with the current (shared) reference sets
  for (sp in ken_mess_spp) {
    mess_surfaces_combined_mess[[sp]][[i + 1]] <-
      if (i == 0) mess_iter0[[sp]] else mess(cov_rast_ken_sub, ref_vals_combined_mess[[sp]])
  }

  if (i == n_iter) break

  # Extract MESS values for each species
  mess_values <- lapply(ken_mess_spp, function(sp) {
    values(mess_surfaces_combined_mess[[sp]][[i + 1]]$mess)
  })
  names(mess_values) <- ken_mess_spp

  # For each cell, find the minimum MESS across all species
  # (the least similar environment to any species' observations)
  min_mess_across_species <- do.call(pmin, mess_values)

  # Select the cell with the overall lowest MESS
  min_idx <- which.min(min_mess_across_species)

  # Get the cell centroid (use first species' raster as template)
  centroid <- xyFromCell(mess_surfaces_combined_mess[[ken_mess_spp[1]]][[i + 1]]$mess, min_idx)
  lon <- centroid[1, "x"]
  lat <- centroid[1, "y"]

  # Determine which species has the minimum MESS at this cell (for tracking)
  mess_at_cell <- sapply(mess_values, \(v) v[min_idx])
  source_species <- names(which.min(mess_at_cell))

  recorded_pts_combined_mess <- rbind(
    recorded_pts_combined_mess,
    data.frame(
      longitude = lon,
      latitude = lat,
      source_species = source_species,
      min_mess_across_spp = min_mess_across_species[min_idx]
    )
  )

  # Add the point to ALL species' reference sets
  new_pt <- vect(data.frame(x = lon, y = lat), geom = c("x", "y"), crs = crs(cov_rast_ken_sub))
  new_vals <- terra::extract(cov_rast_ken_sub, new_pt, ID = FALSE)
  for (sp in ken_mess_spp) {
    ref_vals_combined_mess[[sp]] <- rbind(ref_vals_combined_mess[[sp]], new_vals)
  }
}

mess_results_combined_mess <- setNames(
  lapply(ken_mess_spp, \(sp) list(
    mess_surfaces   = mess_surfaces_combined_mess[[sp]],
    recorded_points = recorded_pts_combined_mess
  )),
  ken_mess_spp
)
```

## Absolute sum of positive and negative cells per iteration (combined MESS-only)

```{r}
#| label: mess-cell-sums-combined-mess
#| fig-width: 12
#| fig-height: 4

cell_sums_combined_mess <- lapply(ken_mess_spp, function(sp) {
  surfaces <- mess_results_combined_mess[[sp]]$mess_surfaces
  data.frame(
    species   = sp,
    iteration = seq_along(surfaces) - 1,
    positive  = sapply(surfaces, function(m) {
      v <- values(m$mess); sum(v[v > 0], na.rm = TRUE)
    }),
    negative  = sapply(surfaces, function(m) {
      v <- values(m$mess); sum(abs(v[v < 0]), na.rm = TRUE)
    })
  )
}) |>
  bind_rows() |>
  pivot_longer(c(positive, negative), names_to = "sign", values_to = "abs_sum")

selected_mess_combined_mess <- lapply(ken_mess_spp, function(sp) {
  surfaces <- mess_results_combined_mess[[sp]]$mess_surfaces
  pts <- mess_results_combined_mess[[sp]]$recorded_points
  data.frame(
    species    = sp,
    iteration  = seq_len(n_iter) - 1,
    mess_value = sapply(seq_len(n_iter), function(j) {
      pt_coord <- as.matrix(pts[j, c("longitude", "latitude")])
      cell_idx <- cellFromXY(surfaces[[j]]$mess, pt_coord)
      values(surfaces[[j]]$mess)[cell_idx]
    })
  )
}) |>
  bind_rows()

plot_data_combined_mess <- bind_rows(
  cell_sums_combined_mess |>
    transmute(species, iteration, series = sign, value = abs_sum,
              panel = "Absolute sum of MESS cells"),
  selected_mess_combined_mess |>
    transmute(species, iteration, series = "selected", value = mess_value,
              panel = "Selected point MESS value")
) |>
  mutate(panel = factor(
    panel,
    levels = c("Absolute sum of MESS cells", "Selected point MESS value")
  ))

ggplot(plot_data_combined_mess, aes(iteration, value, colour = series)) +
  geom_line(data = ~ filter(.x, series == "selected")) +
  geom_point() +
  facet_grid(panel ~ species, scales = "free_y", switch = "y") +
  scale_colour_manual(
    values = c(negative = col_neg, positive = col_pos, selected = "grey40"),
    labels = c(
      negative = "|negative| (novel)",
      positive = "positive (similar)",
      selected = "selected point MESS"
    ),
    breaks = c("negative", "positive", "selected")
  ) +
  labs(x = "Iteration", y = NULL, colour = NULL) +
  theme(strip.placement = "outside")
```

## Absolute sum table (combined MESS-only)

```{r}
#| label: mess-cell-sums-table-combined-mess

cell_sums_combined_mess |>
  pivot_wider(names_from = sign, values_from = abs_sum) |>
  left_join(selected_mess_combined_mess, by = c("species", "iteration")) |>
  arrange(species, iteration) |>
  group_by(species) |>
  mutate(delta_negative = negative - lag(negative)) |>
  ungroup() |>
  knitr::kable(
    digits    = c(0, 0, 0, 0, 2, 0),
    col.names = c("Species", "Iteration", "positive (similar)", "|negative| (novel)", "Selected point MESS", "Δ|negative|")
  )
```

## MESS surface evolution by species (combined MESS-only)

```{r}
#| label: plot-mess-combined-mess
#| fig-width: 10
#| fig-height: 11

clamp_absmax_combined_mess <- max(abs(unlist(lapply(ken_mess_spp, function(sp) {
  lapply(mess_results_combined_mess[[sp]]$mess_surfaces, function(m) {
    quantile(values(m$mess), c(0.01, 0.99), na.rm = TRUE)
  })
}))))

mess_range_combined_mess <- c(-clamp_absmax_combined_mess, clamp_absmax_combined_mess)

for (sp in ken_mess_spp) {
  pts      <- mess_results_combined_mess[[sp]]$recorded_points
  surfaces <- mess_results_combined_mess[[sp]]$mess_surfaces

  layout(layout_mat, heights = c(1, 1, 1, 0.4))
  par(mar = c(1, 1, 2, 0.5), oma = c(0, 0, 3, 0))

  for (j in seq_len(9)) {
    iter <- j - 1
    plot(
      surfaces[[j]]$mess,
      main   = paste0("iter ", iter),
      col    = divergent_pal,
      range  = mess_range_combined_mess,
      legend = FALSE,
      axes   = FALSE
    )
    lines(ken)

    if (iter >= 1) {
      pts_so_far <- pts[seq_len(iter), , drop = FALSE]
      pts_vect   <- vect(pts_so_far, geom = c("longitude", "latitude"), crs = crs(cov_rast_ken_sub))
      points(pts_vect, pch = 19, col = "goldenrod", cex = 2)
    }
  }

  par(mar = c(2.5, 3, 0.5, 3))
  legend_vals <- seq(mess_range_combined_mess[1], mess_range_combined_mess[2], length.out = length(divergent_pal))
  image(
    x = legend_vals, y = 1, z = matrix(legend_vals, ncol = 1),
    col = divergent_pal, axes = FALSE, xlab = "MESS (clamped to 1st/99th pctile)", ylab = ""
  )
  axis(1)

  mtext(bquote(italic(.(sp)) * " (combined MESS-only)"), outer = TRUE, cex = 1.4, font = 2)
}
```




## CV-weighted MESS point selection

Here the species-by-species iterative search is repeated, but instead of selecting
the cell with the lowest MESS value, we select based on a weighted combination:
`weight = MESS × CV`, where CV is the coefficient of variation in the covariates
at each pixel (high CV = high covariate diversity, worth exploring).

```{r}
#| label: mess-iteration-cv-weighted

n_iter <- 20
mess_results_cv <- list()

for (sp in ken_mess_spp) {

  # Start from the species observation points
  ref_vals <- ref_vals_0[[sp]]

  recorded_pts_cv <- data.frame(longitude = numeric(0), latitude = numeric(0))
  mess_surfaces_cv <- vector("list", n_iter + 1)

  for (i in 0:n_iter) {

    # Reuse the pre-computed iteration 0 surface; otherwise recompute
    m <- if (i == 0) mess_iter0[[sp]] else mess(cov_rast_ken_sub, ref_vals)
    mess_surfaces_cv[[i + 1]] <- m

    # Final surface computed; no further points to add
    if (i == n_iter) break

    # Weight by CV: low MESS (novel) × high CV (diverse covariates)
    # Both are masked to Kenya; multiply MESS by CV to get the weighted surface
    weighted <- m$mess * cv[[sp]]

    # Select cell with lowest weighted value (most novel and diverse)
    min_cell <- which.min(values(weighted))
    centroid  <- xyFromCell(weighted, min_cell)
    lon <- centroid[1, "x"]
    lat <- centroid[1, "y"]

    recorded_pts_cv <- rbind(
      recorded_pts_cv,
      data.frame(longitude = lon, latitude = lat)
    )

    # Extract env values at new point and add to reference set
    new_pt   <- vect(data.frame(x = lon, y = lat), geom = c("x", "y"), crs = crs(cov_rast_ken_sub))
    new_vals <- terra::extract(cov_rast_ken_sub, new_pt, ID = FALSE)
    ref_vals <- rbind(ref_vals, new_vals)
  }

  mess_results_cv[[sp]] <- list(
    mess_surfaces = mess_surfaces_cv,
    recorded_points = recorded_pts_cv
  )

  message("Done (CV-weighted): ", sp)
}
```

## Absolute sum of positive and negative cells per iteration (CV-weighted)

```{r}
#| label: mess-cell-sums-cv
#| fig-width: 12
#| fig-height: 4

cell_sums_cv <- lapply(ken_mess_spp, function(sp) {
  surfaces <- mess_results_cv[[sp]]$mess_surfaces
  data.frame(
    species   = sp,
    iteration = seq_along(surfaces) - 1,
    positive  = sapply(surfaces, function(m) {
      v <- values(m$mess); sum(v[v > 0], na.rm = TRUE)
    }),
    negative  = sapply(surfaces, function(m) {
      v <- values(m$mess); sum(abs(v[v < 0]), na.rm = TRUE)
    })
  )
}) |>
  bind_rows() |>
  pivot_longer(c(positive, negative), names_to = "sign", values_to = "abs_sum")

selected_mess_cv <- lapply(ken_mess_spp, function(sp) {
  surfaces <- mess_results_cv[[sp]]$mess_surfaces
  data.frame(
    species    = sp,
    iteration  = seq_len(n_iter) - 1,
    mess_value = sapply(surfaces[seq_len(n_iter)], function(m) {
      min(values(m$mess), na.rm = TRUE)
    })
  )
}) |>
  bind_rows()

plot_data_cv <- bind_rows(
  cell_sums_cv |>
    transmute(species, iteration, series = sign, value = abs_sum,
              panel = "Absolute sum of MESS cells"),
  selected_mess_cv |>
    transmute(species, iteration, series = "selected", value = mess_value,
              panel = "Selected point MESS value")
) |>
  mutate(panel = factor(
    panel,
    levels = c("Absolute sum of MESS cells", "Selected point MESS value")
  ))

ggplot(plot_data_cv, aes(iteration, value, colour = series)) +
  geom_line(data = ~ filter(.x, series == "selected")) +
  geom_point() +
  facet_grid(panel ~ species, scales = "free_y", switch = "y") +
  scale_colour_manual(
    values = c(negative = col_neg, positive = col_pos, selected = "grey40"),
    labels = c(
      negative = "|negative| (novel)",
      positive = "positive (similar)",
      selected = "selected point MESS"
    ),
    breaks = c("negative", "positive", "selected")
  ) +
  labs(x = "Iteration", y = NULL, colour = NULL) +
  theme(strip.placement = "outside")
```

## Absolute sum table (CV-weighted)

```{r}
#| label: mess-cell-sums-table-cv

cell_sums_cv |>
  pivot_wider(names_from = sign, values_from = abs_sum) |>
  left_join(selected_mess_cv, by = c("species", "iteration")) |>
  arrange(species, iteration) |>
  group_by(species) |>
  mutate(delta_negative = negative - lag(negative)) |>
  ungroup() |>
  knitr::kable(
    digits    = c(0, 0, 0, 0, 2, 0),
    col.names = c("Species", "Iteration", "positive (similar)", "|negative| (novel)", "Selected point MESS", "Δ|negative|")
  )
```

## MESS surface evolution by species (CV-weighted)

```{r}
#| label: plot-mess-cv
#| fig-width: 10
#| fig-height: 11

clamp_absmax_cv <- max(abs(unlist(lapply(ken_mess_spp, function(sp) {
  lapply(mess_results_cv[[sp]]$mess_surfaces, function(m) {
    quantile(values(m$mess), c(0.01, 0.99), na.rm = TRUE)
  })
}))))

mess_range_cv <- c(-clamp_absmax_cv, clamp_absmax_cv)

for (sp in ken_mess_spp) {
  pts      <- mess_results_cv[[sp]]$recorded_points
  surfaces <- mess_results_cv[[sp]]$mess_surfaces

  layout(layout_mat, heights = c(1, 1, 1, 0.4))
  par(mar = c(1, 1, 2, 0.5), oma = c(0, 0, 3, 0))

  for (j in seq_len(9)) {
    iter <- j - 1
    plot(
      surfaces[[j]]$mess,
      main   = paste0("iter ", iter),
      col    = divergent_pal,
      range  = mess_range_cv,
      legend = FALSE,
      axes   = FALSE
    )
    lines(ken)

    if (iter >= 1) {
      pts_so_far <- pts[seq_len(iter), , drop = FALSE]
      pts_vect   <- vect(pts_so_far, geom = c("longitude", "latitude"), crs = crs(cov_rast_ken_sub))
      points(pts_vect, pch = 19, col = "goldenrod", cex = 2)
    }
  }

  par(mar = c(2.5, 3, 0.5, 3))
  legend_vals <- seq(mess_range_cv[1], mess_range_cv[2], length.out = length(divergent_pal))
  image(
    x = legend_vals, y = 1, z = matrix(legend_vals, ncol = 1),
    col = divergent_pal, axes = FALSE, xlab = "MESS (clamped to 1st/99th pctile)", ylab = ""
  )
  axis(1)

  mtext(bquote(italic(.(sp)) * " (CV-weighted)"), outer = TRUE, cex = 1.4, font = 2)
}
```

# Combined CV-weighted MESS point selection (all species)

Here the iteration selects a **single location per iteration** from across all
three species by computing `MESS × CV` for each species and cell, then picking
the cell (from any species) with the *lowest* weighted value. This matches the
species-by-species selection direction: because MESS is strongly negative in novel
areas, `MESS × CV` is most negative where a cell is both novel and high-CV, so the
minimum identifies the most novel-and-diverse cell. This location is then added to
all three species' reference sets, so all species evolve together around shared
points.

```{r}
#| label: mess-iteration-combined-cv

ref_vals_combined <- ref_vals_0
recorded_pts_combined <- data.frame(longitude = numeric(0), latitude = numeric(0))
mess_surfaces_combined <- setNames(
  lapply(ken_mess_spp, \(sp) vector("list", n_iter + 1)),
  ken_mess_spp
)

for (i in 0:n_iter) {

  # MESS for every species with the current (shared) reference sets
  for (sp in ken_mess_spp) {
    mess_surfaces_combined[[sp]][[i + 1]] <-
      if (i == 0) mess_iter0[[sp]] else mess(cov_rast_ken_sub, ref_vals_combined[[sp]])
  }

  if (i == n_iter) break

  # For each species and cell, compute weight = MESS × CV
  # Then select the cell with the lowest weight across all species (consistent
  # with the species-by-species which.min selection)
  weights <- lapply(ken_mess_spp, function(sp) {
    mess_surfaces_combined[[sp]][[i + 1]]$mess * cv[[sp]]
  })
  names(weights) <- ken_mess_spp

  # Stack all weighted surfaces and find the global minimum
  all_weights <- lapply(weights, \(w) values(w))
  min_idx <- which.min(unlist(all_weights))

  # Identify which species contributed the min and recover the cell index
  cumsum_len <- cumsum(sapply(all_weights, length))
  sp_idx <- sum(min_idx > c(0, cumsum_len[-length(cumsum_len)]))
  sp_name <- ken_mess_spp[sp_idx]
  cell_in_sp <- min_idx - c(0, cumsum_len[-length(cumsum_len)])[sp_idx]

  # Get the cell centroid
  centroid <- xyFromCell(weights[[sp_name]], cell_in_sp)
  lon <- centroid[1, "x"]
  lat <- centroid[1, "y"]

  recorded_pts_combined <- rbind(
    recorded_pts_combined,
    data.frame(longitude = lon, latitude = lat, source_species = sp_name)
  )

  # Add the point to ALL species' reference sets
  new_pt <- vect(data.frame(x = lon, y = lat), geom = c("x", "y"), crs = crs(cov_rast_ken_sub))
  new_vals <- terra::extract(cov_rast_ken_sub, new_pt, ID = FALSE)
  for (sp in ken_mess_spp) {
    ref_vals_combined[[sp]] <- rbind(ref_vals_combined[[sp]], new_vals)
  }
}

mess_results_combined <- setNames(
  lapply(ken_mess_spp, \(sp) list(
    mess_surfaces   = mess_surfaces_combined[[sp]],
    recorded_points = recorded_pts_combined
  )),
  ken_mess_spp
)
```

## Absolute sum of positive and negative cells per iteration (combined CV-weighted)

```{r}
#| label: mess-cell-sums-combined
#| fig-width: 12
#| fig-height: 4

cell_sums_combined <- lapply(ken_mess_spp, function(sp) {
  surfaces <- mess_results_combined[[sp]]$mess_surfaces
  data.frame(
    species   = sp,
    iteration = seq_along(surfaces) - 1,
    positive  = sapply(surfaces, function(m) {
      v <- values(m$mess); sum(v[v > 0], na.rm = TRUE)
    }),
    negative  = sapply(surfaces, function(m) {
      v <- values(m$mess); sum(abs(v[v < 0]), na.rm = TRUE)
    })
  )
}) |>
  bind_rows() |>
  pivot_longer(c(positive, negative), names_to = "sign", values_to = "abs_sum")

selected_mess_combined <- lapply(ken_mess_spp, function(sp) {
  surfaces <- mess_results_combined[[sp]]$mess_surfaces
  pts <- mess_results_combined[[sp]]$recorded_points
  data.frame(
    species    = sp,
    iteration  = seq_len(n_iter) - 1,
    mess_value = sapply(seq_len(n_iter), function(j) {
      pt_coord <- as.matrix(pts[j, c("longitude", "latitude")])
      cell_idx <- cellFromXY(surfaces[[j]]$mess, pt_coord)
      values(surfaces[[j]]$mess)[cell_idx]
    })
  )
}) |>
  bind_rows()

plot_data_combined <- bind_rows(
  cell_sums_combined |>
    transmute(species, iteration, series = sign, value = abs_sum,
              panel = "Absolute sum of MESS cells"),
  selected_mess_combined |>
    transmute(species, iteration, series = "selected", value = mess_value,
              panel = "Selected point MESS value")
) |>
  mutate(panel = factor(
    panel,
    levels = c("Absolute sum of MESS cells", "Selected point MESS value")
  ))

ggplot(plot_data_combined, aes(iteration, value, colour = series)) +
  geom_line(data = ~ filter(.x, series == "selected")) +
  geom_point() +
  facet_grid(panel ~ species, scales = "free_y", switch = "y") +
  scale_colour_manual(
    values = c(negative = col_neg, positive = col_pos, selected = "grey40"),
    labels = c(
      negative = "|negative| (novel)",
      positive = "positive (similar)",
      selected = "selected point MESS"
    ),
    breaks = c("negative", "positive", "selected")
  ) +
  labs(x = "Iteration", y = NULL, colour = NULL) +
  theme(strip.placement = "outside")
```

## Absolute sum table (combined CV-weighted)

```{r}
#| label: mess-cell-sums-table-combined

cell_sums_combined |>
  pivot_wider(names_from = sign, values_from = abs_sum) |>
  left_join(selected_mess_combined, by = c("species", "iteration")) |>
  arrange(species, iteration) |>
  group_by(species) |>
  mutate(delta_negative = negative - lag(negative)) |>
  ungroup() |>
  knitr::kable(
    digits    = c(0, 0, 0, 0, 2, 0),
    col.names = c("Species", "Iteration", "positive (similar)", "|negative| (novel)", "Selected point MESS", "Δ|negative|")
  )
```

## MESS surface evolution by species (combined CV-weighted)

The same shared points (from the combined optimization) are overlaid on each
species' surface, showing how the globally-optimized picks affect each species.

```{r}
#| label: plot-mess-combined
#| fig-width: 10
#| fig-height: 11

clamp_absmax_combined <- max(abs(unlist(lapply(ken_mess_spp, function(sp) {
  lapply(mess_results_combined[[sp]]$mess_surfaces, function(m) {
    quantile(values(m$mess), c(0.01, 0.99), na.rm = TRUE)
  })
}))))

mess_range_combined <- c(-clamp_absmax_combined, clamp_absmax_combined)

for (sp in ken_mess_spp) {
  pts      <- mess_results_combined[[sp]]$recorded_points
  surfaces <- mess_results_combined[[sp]]$mess_surfaces

  layout(layout_mat, heights = c(1, 1, 1, 0.4))
  par(mar = c(1, 1, 2, 0.5), oma = c(0, 0, 3, 0))

  for (j in seq_len(9)) {
    iter <- j - 1
    plot(
      surfaces[[j]]$mess,
      main   = paste0("iter ", iter),
      col    = divergent_pal,
      range  = mess_range_combined,
      legend = FALSE,
      axes   = FALSE
    )
    lines(ken)

    if (iter >= 1) {
      pts_so_far <- pts[seq_len(iter), , drop = FALSE]
      pts_vect   <- vect(pts_so_far, geom = c("longitude", "latitude"), crs = crs(cov_rast_ken_sub))
      points(pts_vect, pch = 19, col = "goldenrod", cex = 2)
    }
  }

  par(mar = c(2.5, 3, 0.5, 3))
  legend_vals <- seq(mess_range_combined[1], mess_range_combined[2], length.out = length(divergent_pal))
  image(
    x = legend_vals, y = 1, z = matrix(legend_vals, ncol = 1),
    col = divergent_pal, axes = FALSE, xlab = "MESS (clamped to 1st/99th pctile)", ylab = ""
  )
  axis(1)

  mtext(bquote(italic(.(sp)) * " (combined CV-weighted)"), outer = TRUE, cex = 1.4, font = 2)
}
```

