---
title: "Kenya MESS"
format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
  pdf:
    toc: true
    toc-depth: 4
---

## Setup

### Data
```{r}
#| label: setup
#| message: false

library(targets)
library(terra)
library(dplyr)
library(tidyr)
library(ggplot2)
library(bssdm)

# spatial data 
tar_load(
  c(
    covariate_rast_5, 
    model_data_spatial,
    ken,
    target_covariate_names,
    africa_mess,
    africa_mess_nosea
  ),
  store = "~/Documents/tki_work/vector_atlas/va_multispecies_sdm/_targets/"
)

# 
cov_rast_ken <- crop(covariate_rast_5, ken) |>
  mask(ken)

cov_rast_ken_sub <- cov_rast_ken[[target_covariate_names]]

# species for this analysis
ken_mess_spp <- c("arabiensis", "gambiae", "funestus")

# coefficient of variation
cv <- rast("~/Documents/tki_work/vector_atlas/va_multispecies_sdm/spartan_model_comparison/m6/m6_mask_p_cv.tif") |> 
  crop(ken) |> 
  mask(ken)

cv <- cv[[ken_mess_spp]]

# Resample CV to match the covariate raster resolution for later weighting
cv <- resample(cv, cov_rast_ken_sub, method = "bilinear")

# Muted divergent shades used throughout (negative/novel vs positive/similar)
col_neg <- "#D6604D"
col_pos <- "#4393C3"
```

### Helper functions

```{r}
#| label: helpers

# Divergent palette (red = novel, white = edge, blue = similar) and the 3x3 +
# legend layout, both used throughout.
divergent_pal <- colorRampPalette(c(col_neg, "white", col_pos))(255)
layout_mat <- rbind(matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE), rep(10, 3))

# Symmetric range clamped to the 1st/99th percentile across a list of rasters,
# so extreme outliers don't wash out the colour variation.
clamp_range <- function(rasters) {
  absmax <- max(abs(unlist(lapply(
    rasters,
    \(r) quantile(values(r), c(0.01, 0.99), na.rm = TRUE)
  ))))
  c(-absmax, absmax)
}

# Flatten all MESS rasters from a per-species results object into one list.
all_mess_rasters <- function(results, spp = ken_mess_spp) {
  do.call(c, lapply(spp, \(sp) lapply(results[[sp]]$mess_surfaces, \(m) m$mess)))
}

# Horizontal colourbar spanning a panel.
colorbar_h <- function(rng, label, pal = divergent_pal) {
  vals <- seq(rng[1], rng[2], length.out = length(pal))
  image(
    x = vals,
    y = 1,
    z = matrix(vals, ncol = 1),
    col = pal,
    axes = FALSE,
    xlab = label,
    ylab = ""
  )
  axis(1)
}

# 3x3 grid of MESS surfaces (default iterations 0-8, or any 9 requested
# iterations) with selected points overlaid and a shared horizontal colourbar
# below.
plot_mess_evolution <- function(surfaces, pts, mess_range, title,
                                 iters = 0:8,
                                 initial = "initial",
                                 cbar_label = "MESS (clamped to 1st/99th pctile)",
                                 pt_cex = 2,
                                 pt_col = "goldenrod",
                                 pt_pal = NULL) {
  stopifnot(length(iters) == 9)
  # When pt_pal is supplied, colour each point by its selection order (first to
  # last selected) using a colour ramp; otherwise fall back to the single pt_col.
  pt_cols <- if (is.null(pt_pal)) NULL else pt_pal(nrow(pts))
  layout(layout_mat, heights = c(1, 1, 1, 0.4))
  par(mar = c(1, 1, 2, 0.5), oma = c(0, 0, 3, 0))
  for (iter in iters) {
    j <- iter + 1
    plot(surfaces[[j]]$mess,
         main = if (iter == 0) initial else paste0("iter ", iter),
         col = divergent_pal, range = mess_range, legend = FALSE, axes = FALSE)
    lines(ken)
    if (iter >= 1) {
      pts_so_far <- pts[seq_len(iter), , drop = FALSE]
      pts_vect <- vect(
        pts_so_far,
        geom = c("longitude", "latitude"),
        crs = crs(cov_rast_ken_sub)
      )
      col_iter <- if (is.null(pt_cols)) pt_col else pt_cols[seq_len(iter)]
      points(pts_vect, pch = 19, col = col_iter, cex = pt_cex)
    }
  }
  par(
    mar = c(2.5, 3, 0.5, 3)
  )
  colorbar_h(
    mess_range,
    cbar_label
  )
  mtext(
    title,
    outer = TRUE,
    cex = 1.4,
    font = 2
  )
}

# Long-format positive/|negative| cell sums per surface; optional species label.
cell_sums_long <- function(surfaces, species = NULL) {
  df <- data.frame(
    iteration = seq_along(surfaces) - 1,
    positive  = sapply(surfaces, \(m) { v <- values(m$mess); sum(v[v > 0], na.rm = TRUE) }),
    negative  = sapply(surfaces, \(m) { v <- values(m$mess); sum(abs(v[v < 0]), na.rm = TRUE) })
  )
  if (!is.null(species)) df <- cbind(species = species, df)
  pivot_longer(df, c(positive, negative), names_to = "sign", values_to = "abs_sum")
}

# MESS value of the most novel cell at each iteration (minimum of the surface).
selected_mess_min <- function(surfaces, n_iter, species = NULL) {
  df <- data.frame(
    iteration  = seq_len(n_iter) - 1,
    mess_value = sapply(surfaces[seq_len(n_iter)], \(m) min(values(m$mess), na.rm = TRUE))
  )
  if (!is.null(species)) df <- cbind(species = species, df)
  df
}

# MESS value at each recorded point (used when points are chosen on a weighted
# surface, so the minimum MESS isn't necessarily the selected cell).
selected_mess_at_point <- function(surfaces, pts, n_iter, species = NULL) {
  df <- data.frame(
    iteration  = seq_len(n_iter) - 1,
    mess_value = sapply(seq_len(n_iter), function(j) {
      idx <- cellFromXY(surfaces[[j]]$mess, as.matrix(pts[j, c("longitude", "latitude")]))
      values(surfaces[[j]]$mess)[idx]
    })
  )
  if (!is.null(species)) df <- cbind(species = species, df)
  df
}

# Shared colour scale for the iteration summary plots.
mess_series_values <- c(negative = col_neg, positive = col_pos,
                        selected = "grey40", delta = "steelblue")
mess_series_labels <- c(negative = "|negative| (novel)", positive = "positive (similar)",
                        selected = "selected point MESS", delta = "Δ|negative|")

# Iteration summary plot: absolute cell sums, selected-point MESS, and optionally
# the change in novel cells. Facets by species (grid) or by panel only (wrap).
mess_summary_plot <- function(cell_sums, selected_mess,
                              include_delta = FALSE, by_species = TRUE) {
  panel_levels <- c("Absolute sum of MESS cells", "Selected point MESS value")
  if (include_delta) panel_levels <- c(panel_levels, "Δ|negative| (novelty reduction)")

  parts <- list(
    cell_sums |> transmute(across(any_of("species")), iteration,
                           series = sign, value = abs_sum,
                           panel = "Absolute sum of MESS cells"),
    selected_mess |> transmute(across(any_of("species")), iteration,
                               series = "selected", value = mess_value,
                               panel = "Selected point MESS value")
  )
  if (include_delta) {
    delta_df <- cell_sums |> pivot_wider(names_from = sign, values_from = abs_sum)
    if (by_species) delta_df <- group_by(delta_df, species)
    delta_df <- delta_df |>
      mutate(delta_negative = negative - lag(negative)) |>
      ungroup() |>
      transmute(across(any_of("species")), iteration, series = "delta",
                value = delta_negative, panel = "Δ|negative| (novelty reduction)")
    parts <- c(parts, list(delta_df))
  }
  pdata <- bind_rows(parts) |> mutate(panel = factor(panel, levels = panel_levels))
  line_series <- if (include_delta) c("selected", "delta") else "selected"

  breaks_used <- intersect(names(mess_series_labels), unique(pdata$series))
  
  p <- ggplot(pdata, aes(iteration, value, colour = series)) +
    geom_line(data = ~ filter(.x, series %in% line_series)) +
    geom_point() +
    scale_colour_manual(
      values = mess_series_values[breaks_used],
      labels = mess_series_labels[breaks_used],
      breaks = breaks_used
    ) +
    labs(x = "Iteration", y = NULL, colour = NULL) +
    theme(strip.placement = "outside")

  if (by_species) {
    p + facet_grid(panel ~ species, scales = "free_y", switch = "y")
  } else {
    p + facet_wrap(~ panel, scales = "free_y", strip.position = "left")
  }
}

# Iteration summary table (kable), with or without a species column.
mess_summary_table <- function(cell_sums, selected_mess, by_species = TRUE) {
  join_by <- if (by_species) c("species", "iteration") else "iteration"
  tab <- cell_sums |>
    pivot_wider(names_from = sign, values_from = abs_sum) |>
    left_join(
      selected_mess,
      by = join_by
    )
  if (by_species) {
    tab <- tab |> arrange(species, iteration) |> group_by(species)
  } else {
    tab <- tab |> arrange(iteration)
  }
  tab <- mutate(tab, delta_negative = negative - lag(negative))
  if (by_species) tab <- ungroup(tab)
  knitr::kable(
    tab,
    digits = if (by_species) c(0, 0, 0, 0, 2, 0) else c(0, 0, 0, 2, 0),
    col.names = c(
      if (by_species) "Species",
      "Iteration",
      "positive (similar)",
      "|negative| (novel)",
      "Selected point MESS",
      "Δ|negative|"
    )
  )
}

# Species-by-species iterative search. Points are chosen on the MESS surface, or
# on weight_fn(mess, sp) when supplied (e.g. MESS x CV), then added to that
# species' reference set before recomputing.
iterate_by_species <- function(spp, ref_vals_0, mess_iter0, n_iter,
                                weight_fn = NULL, msg = "Done") {
  results <- list()
  for (sp in spp) {
    ref_vals <- ref_vals_0[[sp]]
    recorded_pts <- data.frame(longitude = numeric(0), latitude = numeric(0))
    mess_surfaces <- vector("list", n_iter + 1)

    for (i in 0:n_iter) {
      m <- if (i == 0) mess_iter0[[sp]] else mess(cov_rast_ken_sub, ref_vals)
      mess_surfaces[[i + 1]] <- m
      if (i == n_iter) break

      sel_surface <- if (is.null(weight_fn)) m$mess else weight_fn(m$mess, sp)
      centroid <- xyFromCell(sel_surface, which.min(values(sel_surface)))
      lon <- centroid[1, "x"]; lat <- centroid[1, "y"]

      recorded_pts <- rbind(
        recorded_pts,
        data.frame(longitude = lon, latitude = lat)
      )
      new_pt <- vect(
        data.frame(x = lon, y = lat),
        geom = c("x", "y"),
        crs = crs(cov_rast_ken_sub)
      )
      ref_vals <- rbind(
        ref_vals,
        terra::extract(cov_rast_ken_sub, new_pt, ID = FALSE)
      )
    }

    results[[sp]] <- list(mess_surfaces = mess_surfaces, recorded_points = recorded_pts)
    message(msg, ": ", sp)
  }
  results
}


plot_mess_single <- function(
  mess_surface,
  ken,
  mess_range,
  title = "MESS",
  cbar_label = "MESS (clamped to 1st/99th pctile)",
  divergent_pal,
  pts = NULL
) {
  # Create a 2-panel layout: map + colorbar
  layout_single <- matrix(c(1, 2), nrow = 2, byrow = TRUE)
  layout(layout_single, heights = c(4, 0.5))
  
  # Main MESS map
  par(mar = c(1, 1, 2, 0.5), oma = c(0, 0, 2, 0))
  plot(
    mess_surface$mess,
    col = divergent_pal,
    range = mess_range,
    legend = FALSE,
    axes = FALSE,
    main = title
  )
  lines(ken)
  
  # Overlay recorded points if provided
  if (!is.null(pts) && nrow(pts) > 0) {
    pts_vect <- vect(
      pts,
      geom = c("longitude", "latitude"),
      crs = crs(mess_surface$mess)
    )
    points(pts_vect, pch = 19, col = "goldenrod", cex = 2)
  }
  
  # Horizontal colourbar
  par(mar = c(2.5, 3, 0.5, 3))
  colorbar_h(mess_range, cbar_label)
  
  # Reset layout and margins to defaults
  layout(1)
  par(mar = c(5.1, 4.1, 4.1, 2.1), oma = c(0, 0, 0, 0))
  
  invisible(NULL)
}

```

### Look at some of the data 

#### Environmental Covariates

```{r}
#| label: cov-rast-ken
#| fig-width: 12
#| fig-height: 16
par(mfrow = c (5, 2))
for(i in 1:10) {
  plot(
    x = cov_rast_ken_sub[[i]],
    legend = FALSE,
    axes = FALSE,
    main = names(cov_rast_ken_sub)[i]
  )
  #lines(ken)
}
```

#### Existing vector atlas sample points

```{r}
#| label: point-data

model_data_vect <- vect(
  model_data_spatial |>
    select(
      species,
      longitude,
      latitude,
      count,
      presence,
      data_type
    ) |>
    filter(species %in% ken_mess_spp),
  geom = c("longitude", "latitude"),
  crs = crs(ken)
)


ken_points <- model_data_vect[ken, ]

par(mfrow = c(1, 3), mar = c(2, 2, 2, 1))

for (sp in ken_mess_spp) {
  sp_pts <- ken_points[ken_points$species == sp, ]
  plot(ken)
  title(main = bquote(italic(.(sp))))
  points(sp_pts, pch = 16, col = col_pos, cex = 0.8)
}
```

#### Distribution uncertainty - coefficient of variation

```{r}
#| label: cv-plot
#| fig-width: 12
#| fig-height: 5

# Calculate shared range across all species (symmetric for comparison)
cv_range <- range(
  unlist(
    lapply(
      ken_mess_spp,
      function(sp) {
        quantile(values(cv[[sp]]), c(0.01, 0.99), na.rm = TRUE)
      }
    )
  ),
  na.rm = TRUE
)

# Layout: three map panels above one shared colorbar
layout(rbind(1:3, c(4, 4, 4)), heights = c(1, 0.28))
par(mar = c(2, 2, 3, 1))

for (sp in ken_mess_spp) {
  plot(
    cv[[sp]],
    col = map.pal("magma", 100) |>
      rev(x = _),
    range = cv_range,
    legend = FALSE,
    axes = FALSE
  )
  title(main = bquote(italic(.(sp))))
  lines(ken)
}

# Shared colorbar
par(mar = c(3, 6, 0.5, 6))
magma_pal <- map.pal("magma", 100)
cv_vals <- seq(cv_range[1], cv_range[2], length.out = length(magma_pal))
image(
  x = cv_vals,
  y = 1,
  z = matrix(cv_vals, ncol = 1),
  col = magma_pal,
  axes = FALSE,
  xlab = "Coefficient of Variation (CV)",
  ylab = ""
)
axis(1)
```


## Multivariate Environmental Similarity Surface

The Multivariate Environmental Similarity Surface (MESS) is a method for quantifying how similar environmental conditions at each location are to the reference environmental space defined by observation points. For each pixel in a map, MESS compares its covariate values to the range and distribution of values observed in the reference set (e.g., species occurrences or survey points).

- **Positive MESS values** (blue in maps): The environment is similar to the reference set. Higher values indicate greater similarity—the location falls well within the observed covariate ranges.
- **Negative MESS values** (red in maps): The environment is novel or extrapolated. The location falls *outside* the range of at least one covariate observed in the reference set, indicating potential extrapolation risk for models trained on the reference data.
- **Zero or near-zero**: (white in maps) The location is at or near the edge of the observed environmental space.


### Africa-wide MESS overview

Here we visualize the complete Africa-wide MESS surface with Kenya's boundary overlaid. This shows the environmental novelty landscape across the continent, with Kenya highlighted for reference.

```{r}
#| label: plot-africa-mess-overview
#| fig-width: 11
#| fig-height: 7

# Clamp to 1st/99th percentile to avoid extreme outliers washing out colors
africa_mess_vals <- values(africa_mess$mess)
africa_mess_clamp <- max(abs(quantile(africa_mess_vals, c(0.01, 0.99), na.rm = TRUE)))
africa_mess_range_plot <- c(-africa_mess_clamp, africa_mess_clamp)

# Divergent palette (same as used elsewhere)
divergent_pal <- colorRampPalette(c(col_neg, "white", col_pos))(255)

# Layout: map on left, colorbar on right
layout(
  cbind(1, 2),
  widths = c(1, 0.08)
)

# Map panel
par(
  mar = c(0, 0, 0, 0)
)
plot(
  africa_mess$mess,
  col = divergent_pal,
  range = africa_mess_range_plot,
  axes = FALSE,
  legend = FALSE
)

# Overlay Kenya boundary in black
lines(ken, col = "black", lwd = 2)

# Vertical colorbar panel
par(
  mar = c(3, 1, 3, 1)
)
legend_vals <- seq(
  africa_mess_range_plot[1],
  africa_mess_range_plot[2],
  length.out = length(divergent_pal)
)
image(
  x = 1,
  y = legend_vals,
  z = matrix(legend_vals, nrow = 1),
  col = divergent_pal,
  axes = FALSE,
  xlab = "",
  ylab = "MESS"
)
axis(
  2,
  cex.axis = 0.75,
  las = 1
)
```

###  MESS by species

Here MESS surfaces are calculated for each species separately using their respective observation points as the reference set:

```{r}
#| label: mess-iter0

# Reference env values per species (observation points only)
ref_vals_0 <- lapply(
  ken_mess_spp,
  function(sp) {
    sp_points <- ken_points[ken_points$species == sp, ]
    terra::extract(cov_rast_ken_sub, sp_points, ID = FALSE) |>
      na.omit()
  }
)
names(ref_vals_0) <- ken_mess_spp



# Iteration 0 MESS surface per species
mess_iter0 <- lapply(
  ref_vals_0,
  function(rv) mess(cov_rast_ken_sub, rv)
)
names(mess_iter0) <- ken_mess_spp

# extract just the mess rasters
iter0 <- lapply(mess_iter0, \(m) m$mess)

# Shared symmetric range, clamped to the 1st/99th percentile so extreme outliers
# don't wash out the colour variation. Zero maps to white for comparability.
iter0_range <- clamp_range(iter0)

# Layout: three map panels above one shared horizontal colourbar
layout(rbind(1:3, c(4, 4, 4)), heights = c(1, 0.28))
par(mar = c(2, 2, 3, 1))
for (sp in ken_mess_spp) {
  plot(
    iter0[[sp]],
    col = divergent_pal,
    range = iter0_range,
    legend = FALSE,
    axes = FALSE
  )
  title(main = bquote(italic(.(sp))))
  lines(ken)
}

# Shared colourbar
par(
  mar = c(3, 6, 0.5, 6)
)
colorbar_h(
  iter0_range,
  "MESS (iteration 0)"
)
```

#### Most Dissimilar Variable (MOD)

While MESS identifies *whether* a location is novel by finding the covariate that most deviates from the reference range, the Most Dissimilar variable (MOD) identifies *which* covariate is most responsible for that novelty. For each pixel, MOD shows the index of the covariate with the maximum standardized deviation from its reference range—in other words, the covariate most "out of place" at that location.

```{r}
#| label: plot-iter0-mod
#| fig-width: 12
#| fig-height: 5

# Extract MOD (most dissimilar variable) for each species
# MOD is the index of the variable causing the most dissimilarity
iter0_mod <- lapply(
  mess_iter0,
  \(m) m$mod
)
names(iter0_mod) <- ken_mess_spp

# Define color palette matching each covariate
mod_pal <- c(
  "#2ECC21", "#FF4081", "#F8F32B", "#A8DADC", "#457B9D",
  "#1D3557", "#FF5733", "#F1C40F", "#6C757D", "#C4E538"
)

n_covariates <- nlyr(cov_rast_ken_sub)

# Layout: three map panels above one shared legend panel
layout(rbind(1:3, c(4, 4, 4)), heights = c(1, 0.5))
par(mar = c(2, 2, 3, 1))

for (sp in ken_mess_spp) {
  plot(
    iter0_mod[[sp]],
    col = mod_pal,
    legend = FALSE,
    axes = FALSE,
    type = "classes"
  )
  title(main = bquote(italic(.(sp))))
  lines(ken)
}

# Create a legible legend showing covariate names and colors
par(mar = c(1, 1, 1, 1))
plot(
  NA,
  xlim = c(0, 10),
  ylim = c(0, 10),
  axes = FALSE,
  xlab = "",
  ylab = ""
)

# Arrange legend in two columns for readability
col_1_idx <- seq_len(5)
col_2_idx <- seq(6, 10)
y_pos_1 <- seq(9, 1, by = -1.6)
y_pos_2 <- seq(9, 1, by = -1.6)

# First column (covariates 1-5)
for (i in seq_along(col_1_idx)) {
  cov_idx <- col_1_idx[i]
  rect(
    0.2,
    y_pos_1[i] - 0.4,
    0.8,
    y_pos_1[i] + 0.4,
    col = mod_pal[cov_idx],
    border = "black",
    lwd = 2
  )
  text(
    1.5,
    y_pos_1[i],
    paste0(cov_idx, ". ", target_covariate_names[cov_idx]),
    cex = 0.95,
    adj = c(0, 0.5),
    font = 1
  )
}

# Second column (covariates 6-10)
for (i in seq_along(col_2_idx)) {
  cov_idx <- col_2_idx[i]
  rect(
    5.2,
    y_pos_2[i] - 0.4,
    5.8,
    y_pos_2[i] + 0.4,
    col = mod_pal[cov_idx],
    border = "black",
    lwd = 2
  )
  text(
    6.5,
    y_pos_2[i],
    paste0(cov_idx, ". ", target_covariate_names[cov_idx]),
    cex = 0.95,
    adj = c(0, 0.5),
    font = 1
  )
}
```

#### Pairwise MESS differences between species

The MESS analyses for each species all look simlar. Let's check on that.

Each panel maps the cell-by-cell difference in MESS between pairs of
species.

```{r}
#| label: plot-iter0-diff
#| fig-width: 12
#| fig-height: 5

iter0_diffs <- list(
  "arabiensis - gambiae"  = iter0$arabiensis - iter0$gambiae,
  "arabiensis - funestus" = iter0$arabiensis - iter0$funestus,
  "gambiae - funestus"    = iter0$gambiae    - iter0$funestus
)

diff_absmax <- max(abs(unlist(lapply(iter0_diffs, \(r) minmax(r)))))
diff_range  <- c(-diff_absmax, diff_absmax)

# Distinct divergent palette (orange-teal) for the difference maps, symmetric
# around zero so no difference maps to white.
diff_pal <- colorRampPalette(c("#E08214", "white", "#80CDC1"))(255)

# Layout: three map panels above one shared horizontal colourbar
layout(rbind(1:3, c(4, 4, 4)), heights = c(1, 0.28))
par(mar = c(2, 2, 3, 1))
for (nm in names(iter0_diffs)) {
  plot(
    iter0_diffs[[nm]],
    col = diff_pal,
    range = diff_range,
    legend = FALSE,
    axes = FALSE
  )
  # Parse the pair names to italicize species names
  pair_parts <- strsplit(nm, " - ")[[1]]
  title(main = bquote(italic(.(pair_parts[1])) * " - " * italic(.(pair_parts[2]))))
  lines(ken)
}

# Shared colourbar
par(mar = c(3, 6, 0.5, 6))
colorbar_h(
  diff_range,
  "Difference in iteration 0 MESS",
  pal = diff_pal
)
```

#### Novel (negative) regions -- are they similar across species?

Mask each surface to keep only the negative (novel / extrapolation)
cells, then compare where the novel regions fall across species.

```{r}
#| label: iter0-negative-similarity

# Keep only negative cells (novel areas)
iter0_neg <- lapply(iter0, \(r) { r[r >= 0] <- NA; r })

# Which cells are novel (negative) per species?
is_neg <- lapply(iter0, \(r) values(r) < 0)

# Pairwise agreement in the *location* of novel cells (Jaccard-style overlap)
overlap <- function(a, b) {
  both <- a & b
  either <- a | b
  sum(both, na.rm = TRUE) / sum(either, na.rm = TRUE)
}

cat(
  "Overlap of novel locations (intersection / union):\n",
  "arabiensis vs gambiae: ",
  round(overlap(is_neg$arabiensis, is_neg$gambiae), 3),
  "\n",
  "arabiensis vs funestus:",
  round(overlap(is_neg$arabiensis, is_neg$funestus), 3),
  "\n",
  "gambiae vs funestus:   ",
  round(overlap(is_neg$gambiae, is_neg$funestus), 3),
  "\n"
)

# Correlation of the negative MESS values where both species are novel
neg_cor <- function(a, b) {
  va <- values(a)[, 1]; vb <- values(b)[, 1]
  ok <- !is.na(va) & !is.na(vb)
  if (sum(ok) < 2) return(NA_real_)
  cor(va[ok], vb[ok])
}

cat(
  "\nCorrelation of negative MESS values where both are novel:\n",
  "arabiensis vs gambiae: ",
  round(neg_cor(iter0_neg$arabiensis, iter0_neg$gambiae), 3),
  "\n",
  "arabiensis vs funestus:",
  round(neg_cor(iter0_neg$arabiensis, iter0_neg$funestus), 3),
  "\n",
  "gambiae vs funestus:   ",
  round(neg_cor(iter0_neg$gambiae, iter0_neg$funestus), 3),
  "\n"
)
```

So novel regions for *arabiensis* and *gambiae* are the same and have strong overlap with *funestus*, while they are the same where they are all 


## CV-weighted iterative selection of distinct cells

This is the point-selection method used from here on. A single shared point is
chosen per iteration — the globally most-dissimilar un-chosen cell across the
three species — and added to every species' reference set, so all species evolve
together around shared points. Two features define it:

1. **Selection is weighted by model uncertainty.** The selection surface is
   `MESS × CV`, so the search favours cells that are both environmentally novel
   (strongly negative MESS) and poorly constrained by the current models (high
   CV). This is set with `use_cv_weight <- TRUE`.
2. **Each grid cell can be chosen at most once.** After a cell is selected it is
   masked out, forcing the search to a *new, distinct* cell every iteration.
   Without this, adding a cell that is already inside the reference envelope
   barely moves its own value, and the loop re-picks the same near-zero cell over
   and over once novel space is used up. Masking lets it keep selecting different
   near-zero cells after novelty is exhausted, rather than collapsing onto one.

**Measured behaviour (this run):** total MESS rises steeply during the novelty
phase, as strongly-negative cells are absorbed and the covariate envelope widens
to cover them. Crucially, gains do **not** stop at novelty exhaustion: even once
every cell sits inside the envelope (`MESS ≥ 0` everywhere), each distinct new
point keeps shifting the reference distribution, pulling more cells toward higher
similarity. The trajectory decelerates but is still climbing at iteration 200, so
the full budget is doing useful work. The climb is strongly upward but not
strictly monotonic — a few picks momentarily lower the total, because the
criterion selects the most dissimilar cell rather than maximising the surface sum
directly. The trajectory plot below shows the real shape.

```{r}
#| label: mess-maximise-run

n_iter_maxmess <- 200
use_cv_weight  <- TRUE   # weight the selection surface by CV (MESS × CV)

ref_vals_maxmess <- ref_vals_0
recorded_pts_maxmess <- data.frame(
  longitude      = numeric(0),
  latitude       = numeric(0),
  source_species = character(0),
  iter_added     = integer(0),
  phase          = character(0),
  sel_value      = numeric(0)
)
mess_surfaces_maxmess <- setNames(
  lapply(ken_mess_spp, \(sp) vector("list", n_iter_maxmess + 1)),
  ken_mess_spp
)

# Total MESS summed over every (unmasked) cell and species, per iteration.
total_mess_trajectory <- rep(NA_real_, n_iter_maxmess + 1)

chosen_cells <- integer(0)          # shared-grid cell indices already selected
novelty_exhausted <- FALSE
novelty_exhaustion_iter <- NA_integer_

for (i in 0:n_iter_maxmess) {

  # (Re)compute MESS for every species with the current shared reference sets.
  for (sp in ken_mess_spp) {
    mess_surfaces_maxmess[[sp]][[i + 1]] <-
      if (i == 0) mess_iter0[[sp]] else mess(cov_rast_ken_sub, ref_vals_maxmess[[sp]])
  }

  # Objective: total MESS over the whole surface (all cells, all species).
  total_mess_trajectory[i + 1] <- sum(
    sapply(ken_mess_spp, \(sp) sum(values(mess_surfaces_maxmess[[sp]][[i + 1]]$mess), na.rm = TRUE))
  )

  if (i == n_iter_maxmess) break

  # Per-species selection surface: raw MESS, optionally CV-weighted.
  sel_cols <- lapply(ken_mess_spp, function(sp) {
    m <- mess_surfaces_maxmess[[sp]][[i + 1]]$mess
    if (use_cv_weight) values(m * cv[[sp]])[, 1] else values(m)[, 1]
  })
  names(sel_cols) <- ken_mess_spp

  # For each cell, the most-dissimilar species drives its selection value.
  cell_min <- do.call(pmin, c(sel_cols, na.rm = TRUE))

  # Exclude already-chosen cells so every pick is a distinct location.
  cell_min[chosen_cells] <- NA

  if (all(is.na(cell_min))) {
    message("All selectable cells chosen; stopping at iteration ", i)
    break
  }

  sel_cell  <- which.min(cell_min)      # global most-dissimilar un-chosen cell
  sel_value <- cell_min[sel_cell]
  sp_name   <- ken_mess_spp[which.min(sapply(sel_cols, \(v) v[sel_cell]))]

  # Novelty exhaustion: no negative MESS remains anywhere (bookkeeping only;
  # the selection criterion does not change).
  if (!novelty_exhausted) {
    global_min_mess <- min(
      sapply(ken_mess_spp, \(sp) min(values(mess_surfaces_maxmess[[sp]][[i + 1]]$mess), na.rm = TRUE))
    )
    if (global_min_mess >= 0) {
      novelty_exhausted <- TRUE
      novelty_exhaustion_iter <- i
      message("Novelty exhausted at iteration ", i, ": no negative MESS cells remain.")
    }
  }
  phase_label <- if (novelty_exhausted) "coverage" else "novelty"

  centroid <- xyFromCell(mess_surfaces_maxmess[[sp_name]][[i + 1]]$mess, sel_cell)
  lon <- centroid[1, "x"]
  lat <- centroid[1, "y"]

  recorded_pts_maxmess <- rbind(
    recorded_pts_maxmess,
    data.frame(
      longitude      = lon,
      latitude       = lat,
      source_species = sp_name,
      iter_added     = i,
      phase          = phase_label,
      sel_value      = sel_value
    )
  )
  chosen_cells <- c(chosen_cells, sel_cell)

  # Add the selected point to ALL species' reference sets.
  new_pt <- vect(
    data.frame(x = lon, y = lat),
    geom = c("x", "y"),
    crs = crs(cov_rast_ken_sub)
  )
  new_vals <- terra::extract(cov_rast_ken_sub, new_pt, ID = FALSE)
  for (sp in ken_mess_spp) {
    ref_vals_maxmess[[sp]] <- rbind(ref_vals_maxmess[[sp]], new_vals)
  }
}

# Assemble a results object compatible with the existing summary helpers.
mess_results_maxmess <- setNames(
  lapply(ken_mess_spp, \(sp) list(
    mess_surfaces   = mess_surfaces_maxmess[[sp]],
    recorded_points = recorded_pts_maxmess
  )),
  ken_mess_spp
)

cat("Distinct points selected: ", nrow(recorded_pts_maxmess), "\n", sep = "")
cat("Unique locations:         ",
    nrow(dplyr::distinct(recorded_pts_maxmess, longitude, latitude)), "\n", sep = "")
if (!is.na(novelty_exhaustion_iter)) {
  cat("Novelty exhausted at iteration ", novelty_exhaustion_iter, "\n", sep = "")
} else {
  cat("Novelty not exhausted within ", n_iter_maxmess, " iterations\n", sep = "")
}
cat("Total MESS start -> end: ",
    round(total_mess_trajectory[1]), " -> ",
    round(total_mess_trajectory[length(total_mess_trajectory)]), "\n", sep = "")
```

### Total MESS over the surface across iterations

The key check: does the sum of MESS over the whole surface actually increase as
points are added, and where does it level off?

```{r}
#| label: mess-maximise-trajectory
#| fig-width: 9
#| fig-height: 5

traj_df <- data.frame(
  iteration  = seq_along(total_mess_trajectory) - 1,
  total_mess = total_mess_trajectory
)

p_traj <- ggplot(traj_df, aes(iteration, total_mess)) +
  geom_line(colour = col_pos, linewidth = 0.8) +
  geom_point(colour = col_pos, size = 1) +
  labs(
    x = "Iteration",
    y = "Total MESS (summed over all cells and species)",
    title = "Total surface MESS as distinct points are added"
  ) +
  theme_minimal()

if (!is.na(novelty_exhaustion_iter)) {
  p_traj <- p_traj +
    geom_vline(xintercept = novelty_exhaustion_iter, linetype = "dashed",
               colour = col_neg) +
    annotate("text", x = novelty_exhaustion_iter, y = min(total_mess_trajectory, na.rm = TRUE),
             label = "novelty exhausted", hjust = -0.05, vjust = 0,
             colour = col_neg, size = 3)
}

p_traj
```

### Positive / |negative| decomposition (CV-weighted run)

```{r}
#| label: mess-maximise-summary
#| fig-width: 12
#| fig-height: 4

cell_sums_maxmess <- lapply(
  ken_mess_spp,
  \(sp) cell_sums_long(mess_results_maxmess[[sp]]$mess_surfaces, sp)
) |>
  bind_rows()

selected_mess_maxmess <- lapply(
  ken_mess_spp,
  \(sp) selected_mess_at_point(
    mess_results_maxmess[[sp]]$mess_surfaces,
    mess_results_maxmess[[sp]]$recorded_points,
    nrow(mess_results_maxmess[[sp]]$recorded_points),
    sp
  )
) |>
  bind_rows()

mess_summary_plot(cell_sums_maxmess, selected_mess_maxmess)
```

### Selected points by phase (CV-weighted run)

```{r}
#| label: mess-maximise-map
#| fig-width: 7
#| fig-height: 7

plot(ken, main = "Selected points (CV-weighted, distinct cells)")
phase_cols <- c(novelty = col_neg, coverage = col_pos)
pts_vect <- vect(
  recorded_pts_maxmess,
  geom = c("longitude", "latitude"),
  crs = crs(cov_rast_ken_sub)
)
points(pts_vect, pch = 19, cex = 0.9,
       col = phase_cols[recorded_pts_maxmess$phase])
legend("topright", legend = names(phase_cols), pch = 19,
       col = phase_cols, bty = "n", title = "phase")
```

### Export selected points (CV-weighted run)

```{r}
#| label: mess-maximise-export

readr::write_csv(
  recorded_pts_maxmess |> mutate(priority = row_number()),
  "selected_points_maxmess.csv"
)

cat("Points by phase:\n")
recorded_pts_maxmess |>
  group_by(phase) |>
  summarise(count = n()) |>
  print()
```

### MESS surface evolution (CV-weighted run)

A 3x3 evolution grid per species for the CV-weighted distinct-cells run.
Iteration 0, iteration `n_iter_maxmess`, and 7 evenly-spaced iterations in
between span the full 0–200 range in one grid, with the selected points
accumulated so far overlaid. Points are coloured on a viridis ramp from first
(dark) to last (yellow) selected, and drawn small since up to 200 are shown at
once.

```{r}
#| label: plot-mess-maximise-evolution
#| fig-width: 10
#| fig-height: 11

mess_range_maxmess <- clamp_range(all_mess_rasters(mess_results_maxmess))

iters_maxmess_full <- round(seq(0, n_iter_maxmess, length.out = 9))

for (sp in ken_mess_spp) {
  plot_mess_evolution(
    surfaces   = mess_results_maxmess[[sp]]$mess_surfaces,
    pts        = mess_results_maxmess[[sp]]$recorded_points,
    mess_range = mess_range_maxmess,
    title      = bquote(italic(.(sp)) * " (CV-weighted, distinct cells)"),
    iters      = iters_maxmess_full,
    initial    = "iter 0",
    pt_cex     = 1,
    pt_pal     = \(n) hcl.colors(n, "viridis")
  )
}
```

## ExDet surface evolution over the CV-weighted MESS points

Rather than running a separate ExDet-driven search, here we reuse the **same 200
points** selected by the CV-weighted MESS run above (`recorded_pts_maxmess`, in
selection order) and recompute the **Extrapolation Detection** surface
(`bssdm::exdet`) as those points are progressively added to the reference set.
This shows how environmental novelty — resolved into its two ExDet components —
evolves under the MESS-chosen sampling design.

ExDet packs both novelty types into a single value per cell:

- **`exdet < 0` — Type 1 (univariate) novelty:** the cell lies *outside* the
  observed range of at least one covariate; more negative = more novel.
- **`0 ≤ exdet ≤ 1` — within the reference envelope** (no univariate
  extrapolation, inside the combinational Mahalanobis envelope).
- **`exdet > 1` — Type 2 (combinational) novelty:** every covariate is
  individually in range but their *combination* is novel; larger = more novel.

```{r}
#| label: exdet-over-maxmess-setup

# ExDet is coloured with *deterministic discrete classes*, not a continuous
# ramp. A hand-built continuous palette passed to terra as a long `col` vector
# is re-interpolated by terra's continuous renderer, which bleeds the near-zero
# "Type 1" colour into cells with exdet just above 0 (the most *typical* cells).
# Those typical cells multiply as points are added, so the map grows a spurious
# reddish region that looks — misleadingly — like novelty increasing. Fixed
# `breaks` assign each cell to a class by value, with no interpolation, so the
# [0, 1] "similar" band is always grey.
#
# Classes:      exdet < 0  -> Type 1 (univariate), reds, darker = more novel
#          0 <= exdet <= 1 -> similar, grey
#               exdet > 1  -> Type 2 (combinational), purples, darker = more novel
# Note the lower "similar" bound is -1e-9 so that exdet == 0 lands in the grey
# band rather than the reddest class.
exdet_breaks <- c(-Inf, -1, -0.25, -1e-9, 1, 2, 4, Inf)
exdet_cols   <- c("#67001F", "#D6604D", "#FDDBC7",   # Type 1: strong -> weak
                  "#f0f0f0",                          # similar
                  "#E7D4E8", "#9970AB", "#40004B")    # Type 2: weak -> strong
exdet_class_labels <- c(
  "Type 1 (< -1)", "Type 1 (-1 to -0.25)", "Type 1 (-0.25 to 0)",
  "similar (0-1)",
  "Type 2 (1-2)", "Type 2 (2-4)", "Type 2 (> 4)"
)

# Covariate values at the 200 CV-weighted MESS points, in selection order. The
# same shared points are added to every species' reference set.
maxmess_pts_cov <- terra::extract(
  cov_rast_ken_sub,
  vect(recorded_pts_maxmess, geom = c("longitude", "latitude"),
       crs = crs(cov_rast_ken_sub)),
  ID = FALSE
)

# Iterations to display in the evolution grid: iteration 0, iteration 200, and 7
# evenly spaced in between (same grid as the MESS evolution plot).
iters_exdet_full <- round(seq(0, n_iter_maxmess, length.out = 9))

# Recompute the ExDet surface as each MESS point is added, for *every* iteration
# 0..n_iter_maxmess and every species. Per iteration we record the total novelty
# of each type, summed over all cells and species:
#   * Type 1 total: sum of |exdet| where exdet < 0 (univariate extrapolation), and
#   * Type 2 total: sum of (exdet - 1) where exdet > 1 (combinational novelty
#     above the reference envelope).
# Only the 9 grid iterations' surfaces are retained (for the evolution plot); the
# rest are used for the trajectory sums and then discarded to save memory.
exdet_surfaces_maxmess <- setNames(
  lapply(ken_mess_spp, \(sp) vector("list", length(iters_exdet_full))),
  ken_mess_spp
)
exdet_novelty_traj <- data.frame(
  iteration = 0:n_iter_maxmess,
  type1     = NA_real_,
  type2     = NA_real_
)

for (i in 0:n_iter_maxmess) {
  t1 <- 0
  t2 <- 0
  g  <- match(i, iters_exdet_full)   # grid slot, or NA if not a displayed iter
  for (sp in ken_mess_spp) {
    ref <- if (i == 0) {
      ref_vals_0[[sp]]
    } else {
      rbind(ref_vals_0[[sp]], maxmess_pts_cov[seq_len(i), , drop = FALSE])
    }
    ex <- exdet(cov_rast_ken_sub, ref)$exdet
    v  <- values(ex)[, 1]
    t1 <- t1 + sum(abs(v[v < 0]), na.rm = TRUE)
    t2 <- t2 + sum(v[v > 1] - 1,  na.rm = TRUE)
    if (!is.na(g)) exdet_surfaces_maxmess[[sp]][[g]] <- ex
  }
  exdet_novelty_traj$type1[i + 1] <- t1
  exdet_novelty_traj$type2[i + 1] <- t2
}
message("ExDet trajectory + grid surfaces computed over ",
        n_iter_maxmess, " iterations")
```

### ExDet surface evolution grid

A 3x3 grid per species showing the ExDet surface at each displayed iteration,
with the MESS-selected points accumulated so far overlaid (coloured on a viridis
ramp from first selected, dark, to last, yellow). As expected, the Type 1
(red) extrapolation region collapses almost entirely once the first handful of
extreme points are added; a residual Type 2 (purple) combinational-novelty patch
in the highlands persists and slowly shrinks — total novel area only ever
decreases.

```{r}
#| label: plot-exdet-over-maxmess
#| fig-width: 10
#| fig-height: 11

# 3x3 grid of ExDet surfaces (supplied for the 9 displayed iterations) with the
# points selected so far overlaid, coloured by selection order. Colours are
# assigned by fixed `breaks` (discrete classes) rather than a continuous palette,
# so the [0, 1] "similar" band is always grey.
plot_exdet_evolution <- function(surfaces, iters, pts, title,
                                 pt_cex = 1, pt_pal = \(n) hcl.colors(n, "viridis")) {
  stopifnot(length(surfaces) == 9, length(iters) == 9)
  pt_cols <- pt_pal(nrow(pts))
  layout(layout_mat, heights = c(1, 1, 1, 0.5))
  par(mar = c(1, 1, 2, 0.5), oma = c(0, 0, 3, 0))
  for (k in seq_len(9)) {
    iter <- iters[k]
    plot(surfaces[[k]],
         main = if (iter == 0) "iter 0" else paste0("iter ", iter),
         breaks = exdet_breaks, col = exdet_cols, legend = FALSE, axes = FALSE)
    lines(ken)
    if (iter >= 1) {
      pts_so_far <- pts[seq_len(iter), , drop = FALSE]
      pv <- vect(pts_so_far, geom = c("longitude", "latitude"),
                 crs = crs(cov_rast_ken_sub))
      points(pv, pch = 19, col = pt_cols[seq_len(iter)], cex = pt_cex)
    }
  }
  # Discrete class legend spanning the bottom row.
  par(mar = c(0, 1, 0, 1))
  plot.new()
  legend("center", legend = exdet_class_labels, fill = exdet_cols,
         border = NA, horiz = FALSE, ncol = 4, bty = "n",
         title = "ExDet class", cex = 1.1)
  mtext(title, outer = TRUE, cex = 1.4, font = 2)
}

for (sp in ken_mess_spp) {
  plot_exdet_evolution(
    surfaces = exdet_surfaces_maxmess[[sp]],
    iters    = iters_exdet_full,
    pts      = recorded_pts_maxmess,
    title    = bquote(italic(.(sp)) * " — ExDet over CV-weighted MESS points"),
    pt_cex   = 1
  )
}
```

### Total novelty over iterations

Total Type 1 and Type 2 novelty, summed over every cell and species, as the 200
CV-weighted MESS points are added. **Type 1** (univariate extrapolation)
collapses to zero within the first ~20 points: each extreme MESS point widens the
observed range of some covariate, quickly bringing every cell inside the
univariate envelope. **Type 2** (combinational novelty) behaves very differently
— it *rises* at first, because those same extreme points distort the reference
correlation structure and push more cells outside the combinational envelope,
before slowly declining as later points fill the space in. It never reaches zero
within the budget: combinational novelty is much harder to eliminate than
univariate novelty.

```{r}
#| label: exdet-novelty-trajectory
#| fig-width: 9
#| fig-height: 5

exdet_traj_long <- exdet_novelty_traj |>
  pivot_longer(c(type1, type2), names_to = "novelty", values_to = "total") |>
  mutate(novelty = recode(novelty,
    type1 = "Type 1 (univariate)",
    type2 = "Type 2 (combinational)"))

ggplot(exdet_traj_long, aes(iteration, total, colour = novelty)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c(
    "Type 1 (univariate)"    = "#D6604D",
    "Type 2 (combinational)" = "#762A83"
  )) +
  labs(
    x = "Iteration",
    y = "Total novelty (summed over cells and species)",
    colour = NULL,
    title = "ExDet novelty as CV-weighted MESS points are added"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```
